---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
# load tidyverse packages, including ggplot2

# read in the data
gun_deaths <- read.csv("gun_deaths.csv")

head(gun_deaths)
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
deaths_per_month <- gun_deaths %>%
    count(month)%>%
    rename(deaths = n)

deaths_per_month<-transform(deaths_per_month, MonthAbb = month.abb[month])
deaths_per_month$MonthAbb <- factor(deaths_per_month$MonthAbb, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

deaths_per_month
    
```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
ggplot(deaths_per_month, aes(MonthAbb, deaths))+
    geom_bar(stat="identity")+
    labs(x="Month", y="Deaths", title="Deaths Per Month")

deaths_per_month
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
death_intent <- gun_deaths%>%
    count(intent)%>%
    rename(deaths=n)%>%
    arrange(deaths)

ggplot(death_intent, aes(x=reorder(intent, -deaths), y=deaths))+
    geom_bar(stat="identity")+
    labs(x="Intent", y="Deaths", title="Deaths per Intent Cause")

death_intent
    

```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}

sex_age_death <- gun_deaths%>%
    filter(age !="Not Available")%>%
    mutate(age = as.numeric(age))%>%
    select(sex, age)%>%
    group_by(age)%>%
    count(sex)%>%
    arrange(age)%>%
    rename(deaths=n)%>%
    ungroup()

ggplot(sex_age_death, aes(x=sex, y=age, color=sex))+
    geom_boxplot()


sex_age_death%>%
    filter(sex=="F")%>%
    summarize(average_f=mean(age))

```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
gun_deaths%>%
    filter(race=="White", sex=="M", education!="Less than HS", year ==2012)%>%
    count()%>%
    rename(deaths=n)
```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
winter <- c("January, February", "March")
spring <- c("April", "June", "July")
summer <- c("July", "August", "September")
fall <- c("October", "November", "December")
seasons <- c(winter, spring, summer, fall)

season_death <- gun_deaths%>%
    mutate 
    

```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
suicide_or_homicide <- gun_deaths%>%
    select(race, intent)%>%
    filter(intent %in% c("Suicide", "Homicide"), race %in% c("White", "Black", "Hispanic"))%>%
    group_by(race, intent)%>%
    count()%>%
    rename(deaths=n)

suicide_or_homicide

```

(hint maybe looking at percentages would be good)
```{r race_2}
suicide_or_homicide <- gun_deaths%>%
    select(race, intent)%>%
    filter(intent %in% c("Suicide", "Homicide"), race %in% c("White", "Black", "Hispanic"))%>%
    group_by(race, intent)%>%
    count()%>%
    rename(deaths=n)%>%
    ungroup()%>%
    group_by(race)%>%
    mutate(percentage_of_race_death = deaths/sum(deaths)*100)

suicide_or_homicide

```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}
police_deaths <- gun_deaths%>%
    mutate(age = as.numeric(age))%>%
    select(police, age, race)%>%
    group_by(race)%>%
   

ggplot(police_deaths, mapping=aes(x=factor(police), y=age))+
    geom_boxplot()+
    labs(x = "Killed by Police?")


```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

##Unfortunately, I ran out of time to finish the rest of the assignment. I got stuck finding a way to adequately represent the police related deaths to race. My plan moving forward would have been to do some searching on how to do statists tests within R. I was also unable to group the data into months to complete the seasonal data analysis.

```{r stats?}


```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

