---
title: "Exploring Gun Deaths in America"
author: "Ben Wang"
date: "7/27/20"
output: github_document
---

```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)   
library(ggplot2)
library(knitr)
thedata <- read.csv("gun_deaths.csv")
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
gun_deaths_per_month <- thedata %>%
    select(month) %>%
    group_by(month) %>%
    count(month) %>%
    rename(deaths_per_month = n)

kable(gun_deaths_per_month, align = 'c', col.names = c("Month", "Deaths/Month"))
```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
thedata$month <- factor(thedata$month, labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

ggplot(thedata, aes(x = month, fill = month)) +
    geom_bar() + 
    labs(x = "Month", y = "NUmber of Deaths") 
```
## Is there a way to make my axis text smaller so they're not overlapping like this? Also, I'm going to try and make the graph slightly less colorful. As it stands I think it's a bit too flamboyant. 

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
intent_graph <- thedata %>%
    group_by(intent) %>%
    count() %>%
    rename(number_of_deaths = n)
## Going back to give kable another shot, it seems to be working better for me than ggplot (previous figure)
## Also, would you guys happen to know a way to order these rows by ascending/descending n? I'm curious if there's a function to do that automatically without having to manually adjust it ourselves.
kable(intent_graph, align = 'c', col.names = c("Intent", "Number of Deaths"))
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
gun_deaths_sex <- thedata %>%
    select(sex, age) %>%
    group_by(sex)

gun_deaths_sex$age <- as.integer(gun_deaths_sex$age)
 
ggplot(gun_deaths_sex) +
    geom_boxplot(aes(sex, age, fill = sex)) +
    labs(title = "Age of gun death victims sorted by sex", x = "Sex", y = "Age") +
    theme(legend.position = "none")
    
# find the average age of female gun death victims
gun_deaths_average_female <- gun_deaths_sex %>%
    filter(sex == "F") %>%
    summarize(average_age = mean(age))
print(gun_deaths_average_female)
```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
gun_deaths_education <- thedata %>%
    select(sex, education, year, race) %>%
    filter(sex == "M", year == 2012, race == "White", education %in% c('HS/GED', 'Some College', 'BA+')) %>%
    count() 

kable(gun_deaths_education, align = 'c', col.names = c('Number of Deaths'))
```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
## Not really sure how to do this one, do we filter for each individual month then add those numbers up and create another table? 
```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
gun_deaths_race <- thedata %>%
    select(intent, race) %>%
    filter(intent %in% c('Suicide', 'Homicide')) %>%
    filter(race %in% c('White', 'Black', 'Hispanic')) %>%
    count(race)

ggplot(gun_deaths_race, aes(x = gun_deaths_race$race, y = n, fill = intent)) +
    labs(x = "Race", y = "Number of Deaths") 
```
## Seems like I got stuck here and ran out of time this week. When I run the code up to here it gives me an error stating that object 'intent' could not be found? Will come back to this and try to figure it out when I have more time.

(hint maybe looking at percentages would be good)
```{r race_2}

```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}

```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}


```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

