---
title: "Exploring Gun Deaths in America"
author: Sandra McClure
date: July 20th, 2020
output: github_document
---
 
## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(ggplot2)
library(knitr)
gun_deaths <- read.csv( file = "gun_deaths.csv", head = TRUE, sep = ",")

# read in the data

```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
GunDeaths_month <- gun_deaths %>%
    select(month) %>%
    mutate(month=factor(month,labels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"))) %>%
    count(month)
kable(GunDeaths_month)

```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
ggplot(data = GunDeaths_month, aes(x = month, y = n )) + 
    labs(title = "Gun Deaths per Month", x = "Month", y = "Number of Deaths") + 
    geom_bar(stat = "identity")
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
GunDeaths_intent <- gun_deaths %>%
    count(intent)

ggplot(GunDeaths_intent, aes(x = intent, y = n))+
    geom_bar( stat = "identity") +
    labs (Title = "Gun Deaths Assciated With Intent Cause of Death", x = "Intent", y = "Deaths")
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
GunDeaths_age_sex <- gun_deaths %>%
    select(age, sex)
    
ggplot(GunDeaths_age_sex, aes(x = sex, y = age))  +
    geom_boxplot() + 
    labs(title = "Gun Deaths Associated with Age and Sex", x = "Sex", y = "Age")

GunDeaths_age_sex %>% 
    filter(sex =="F") %>%
    summarize(age_ave = mean(age)) %>%
    print()
```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
GunDeaths_white_males__HS_2012 <- gun_deaths %>% 
    select(year, sex, race, education) %>% 
    filter(year == 2012, sex == "M", race == "White",   education != "Less than HS") %>% 
    count()
print(GunDeaths_white_males__HS_2012)
```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
Winter <- gun_deaths %>%
    select(month) %>%
    filter(month %in% c("Jan", "Feb", "Mar"))

Spring <- gun_deaths %>%
    select(month) %>%
    filter(month %in% c("Apr", "May", "Jun"))

Summer <- gun_deaths %>%
    select(month) %>%
    filter(month %in% c("Jul", "Aug", "Sep"))

Fall <- gun_deaths %>%
    select(month) %>%
    filter(month %in% c("Oct", "Nov", "Dec"))

Seasons <- c(Winter, Spring, Summer, Fall)

GunDeaths_season <- gun_deaths %>%
    summarize (winter = sum("Winter"), spring = sum("Spring"), summer = sum("Summer"), fall = sum("Fall")) %>%
    count() %>%
    arrange(desc(Seasons)) %>%
    print()


```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
GunDeaths_Suic_Homi <- gun_deaths %>%
    select (race, intent) %>%
    filter ( race == c("White", "Hispanic", "Black"), intent == c("Suicide", "Homicide")) %>%
    group_by(race, intent)
    count() %>%
    print()
```

(hint maybe looking at percentages would be good)
```{r race_2}
GunDeaths_Suic_Homi <- gun_deaths %>%
    select(race, intent) %>%
    filter(intent %in% c("Suicide", "Homicide"), race %in% c("White", "Black", "Hispanic")) %>%
    group_by(race) %>%
    count() %>%
    mutate(GunDeaths_Suic_Homi_percent = n/sum(n)*100) 

print(GunDeaths_Suic_Homi)
```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}

```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}


```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

