---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(knitr)
library(scales)

# read in the data
gun_deaths<-read_csv("gun_deaths.csv") #col_types="cccdcdcdc"
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
by_month<-gun_deaths%>%
    filter(year==2012)%>% #I'm just working with data from 2012, because I was having trouble making nice looking graphs using data with the other years
    group_by(month)%>%
    summarize(n=n()) #counts the number of rows for each month
kable(by_month)
```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
ggplot(by_month,aes(x=factor(month),y=n))+
    geom_col()+
    labs(x="months",y="number of gun deaths",title="number of gun deaths per month in 2012")+
    scale_x_discrete(labels=c("1"="Jan","2"="Feb","3"="Mar","4"="Apr","5"="May","6"="Jun","7"="Jul","8"="Aug","9"="Sep","10"="Oct","11"="Nov","12"="Dec"))
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
by_intent<-gun_deaths%>%
    group_by(intent)%>%
    summarize(n=n())%>%
    arrange(desc(n)) #Kept this line here, but I don't think it's necessary

ggplot(by_intent,aes(x=factor(intent, level=c("Suicide","Homicide","Accidental","Undetermined","Not Available")),y=n))+
    geom_col()+
    labs(x="type of intent cause of death",y="number of gun deaths",title="number of gun deaths by type of intent cause of death")
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
by_age_sex<-gun_deaths%>%
    arrange(desc(age))%>%
    slice(19:n()) #first 18 rows were age=Not available

by_age_sex$age<-as.numeric(by_age_sex$age) #the age column was read as character, so I used this to change it to numeric to allow for proper plotting
    
ggplot(by_age_sex,aes(x=sex,y=age))+
    geom_boxplot()+
    labs(x="sex",y="age",title="boxplot of ages of gun death victims by sex")

by_age_sex_female<-by_age_sex%>%
    filter(sex=="F")

print(paste("The average age of female gun death victims is ",mean(by_age_sex_female$age)))
```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
#I didn't know how to filter for all the correct education values, so I created separate data frames for each and summed the counts at the end lol
by_hs_wm_ba<-gun_deaths%>%
    filter(year==2012)%>%
    filter(education=="BA+")%>%
    #filter(education=="Some college")%>%
    #filter(education=="HS/GED")%>%
    filter(race=="White")%>%
    filter(sex=="M")%>%
    group_by(sex)%>%
    summarize(n=n())

by_hs_wm_sc<-gun_deaths%>%
    filter(year==2012)%>%
    #filter(education=="BA+")%>%
    filter(education=="Some college")%>%
    #filter(education=="HS/GED")%>%
    filter(race=="White")%>%
    filter(sex=="M")%>%
    group_by(sex)%>%
    summarize(n=n())

by_hs_wm_hs<-gun_deaths%>%
    filter(year==2012)%>%
    #filter(education=="BA+")%>%
    #filter(education=="Some college")%>%
    filter(education=="HS/GED")%>%
    filter(race=="White")%>%
    filter(sex=="M")%>%
    group_by(sex)%>%
    summarize(n=n())

print(paste(sum(by_hs_wm_ba[2:2],by_hs_wm_sc[2:2],by_hs_wm_hs[2:2])," white males with at least a high school education were killed by guns in 2012."))
```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
by_season<-gun_deaths%>%
    mutate(season=month)

by_season$season<-gsub(pattern="10|11|12", replacement = "Fall",x=by_season$season) #need to replace Fall months first to get rid of the numbers that will be replaced later

by_season$season<-gsub(pattern="1|2|3", replacement = "Winter",x=by_season$season)

by_season$season<-gsub(pattern="4|5|6", replacement = "Spring",x=by_season$season)

by_season$season<-gsub(pattern="7|8|9", replacement = "Summer",x=by_season$season)

by_season_count<-by_season%>%
    group_by(season)%>%
    summarize(n=n())

ggplot(by_season_count,aes(x=factor(season,level=c("Winter","Spring","Summer","Fall")),y=n))+
    geom_col()+
    labs(title="barplot of number gun deaths by season",y="number of gun deaths",x="season")

print("Summer has the most gun deaths")
```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
by_intent_race<-gun_deaths%>%
    group_by(intent,race)%>%
    summarize(n=n())

ggplot(by_intent_race,aes(x=race,y=n,fill=intent))+
    geom_col(position="dodge")+
    theme(axis.text.x = element_text(vjust=0.5))+
    labs(title="barplot of gun deaths by race and intent",y="number of gun deaths")+
    scale_x_discrete(labels=wrap_format(20))#to wrap long tick labels into multiple lines
```

(hint maybe looking at percentages would be good)
```{r race_2}
by_intent_race_percent<-gun_deaths%>% #to create percentages
    count(race,intent)%>%
    group_by(race)%>%
    mutate(percent = n/sum(n)*100)%>%
    ungroup()

ggplot(by_intent_race_percent,aes(x=race,y=percent,fill=intent))+
    geom_col(position="dodge")+
    theme(axis.text.x = element_text(vjust=0.5))+
    labs(title="barplot of percentages of gun deaths by intent for different races",y="percentage of total gun deaths")+
    scale_x_discrete(labels=wrap_format(20))

print("Whites who are killed by guns more likely to die because of suicide whereas blacks and hispanics who are killed by guns more likely to die because of homicide.")
```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}
by_age<-gun_deaths%>%
    arrange(desc(age))%>%
    slice(19:n())

by_age$age<-as.numeric(by_age$age)

by_age$police<-gsub(pattern="0",replacement="No police involvement",x=by_age$police)

by_age$police<-gsub(pattern="1",replacement="Police involvement",x=by_age$police)

ggplot(by_age,aes(x=police,y=age))+
    geom_boxplot()+
    labs(title="boxplot of age of gun deaths by police-involvement",x=NULL)

by_race_percent<-gun_deaths%>%
    count(race,police)%>%
    group_by(race)%>%
    mutate(percent = n/sum(n)*100)%>%
    ungroup()%>%
    filter(police==1)

by_race_percent$police<-gsub(pattern="0",replacement="No police involvement",x=by_race_percent$police)

by_race_percent$police<-gsub(pattern="1",replacement="Police involvement",x=by_race_percent$police)

ggplot(by_race_percent,aes(x=race,y=percent,fill=police))+
    geom_col(position="dodge")+
    scale_x_discrete(labels=wrap_format(20))+
    theme(legend.position="none")+
    labs(y="percentage of gun deaths with police involvement",title="barplot of percentage of gun deaths with police involvement by race")
```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}
t.test(by_age$age~by_age$police,mu=0,alt="two.sided",conf=0.95,var.eq=F,paired=F)

```


```{r police_2}
#not sure what this chunk is for
```

```{r stats_2?}
#proportion test to assess police involvement and api vs not api. not sure I did this right, so I didn't do it for the other races.
by_api<-gun_deaths

#simplify race data to api and not api
by_api$race<-gsub(pattern="Black",replacement="Not API",x=by_api$race)
by_api$race<-gsub(pattern="Hispanic",replacement="Not API",x=by_api$race)
by_api$race<-gsub(pattern="Not Availabletive American/Not Availabletive Alaskan",replacement="Not API",x=by_api$race)
by_api$race<-gsub(pattern="White",replacement="Not API",x=by_api$race)

by_api<-by_api%>%
    group_by(race,police)%>%
    summarize(n=n())

#set up matrix for by_api as input for proportion test
api <- matrix(c(by_api$n[1],by_api$n[2],by_api$n[3],by_api$n[4]), ncol=2)
colnames(api) <- c('api','not api')
rownames(api) <- c('no police involvement','police involvement')

prop.test(api)
```


Finally, all together now:
```{r police_3}
#I'm a noob at stats and don't know how to do this.
```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

