---
title: "Exploring Gun Deaths in America"
author: Mingyu Qi
date: 07/27/2020
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(knitr)
library(ggrepel)

# read in the data
gun_deaths <- read.csv("gun_deaths.csv")

# check variables in the data
glimpse(gun_deaths)

# convert age from character to numeric 
gun_deaths$age <- as.numeric(gun_deaths$age)

# clean up the labels of race
gun_deaths$race <- gsub(pattern = "Not Availabletive American/Not Availabletive Alaskan", 
     replacement = "Native American / Native Alaskan",
     x = gun_deaths$race)

# convert race from character to factor
gun_deaths$race <- factor(gun_deaths$race)
summary(gun_deaths$race)
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
gun_deaths_month <- gun_deaths %>%
    group_by(month) %>%
    count(month) %>%
    rename(number = n)

kable(gun_deaths_month, format = "markdown")
```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
# create human-readable label for each month
gun_deaths$month <- factor(gun_deaths$month, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

# Draw the bar chart
ggplot(gun_deaths, aes(x = month, fill = month)) +
    geom_bar() + 
    xlab("\nMonth") + 
    ylab("Number of Death\n") +
    scale_y_continuous(breaks=seq(0, 9000, by = 1000))
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
ggplot(gun_deaths, aes(x=fct_infreq(intent), fill=intent)) +
    geom_bar() +
    xlab("\nType of Intent") + 
    ylab("Number of Death\n") +
    scale_y_continuous(breaks=seq(0, 70000, by = 10000))
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
# calculate average age by sex
avg_age <- gun_deaths %>%
    group_by(sex) %>%
    summarize(mean_age=mean(age, na.rm = TRUE)) 

# round to the nearest tenth
avg_age$mean_age <- round(avg_age$mean_age, digits = 2)

# draw boxplot by sex and print average age
ggplot(gun_deaths, aes(x = age, y = sex)) +
    geom_boxplot() +
    xlab("\nAge") + 
    ylab("Sex\n") +
    scale_x_continuous(breaks=seq(0, 100, by = 10)) + 
    stat_summary(fun = mean, colour = "red", geom = "point", shape = 18, size = 2) +
    geom_text_repel(avg_age, mapping = aes(x = mean_age, y = sex, label = mean_age), inherit.aes = FALSE)
```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
# convert education to a factor 
gun_deaths$education <- factor(gun_deaths$education)

# get a summary of education 
summary(gun_deaths$education)

# use filter to select the records we are interested in
gun_deaths %>%
    filter(race == "White" & education == c("HS/GED", "Some college", "BA+") & year == 2012) %>%
    count()
```
** 6,062 white males with at least a high school education were killed by guns in 2012

## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
# create factor "season" in a new data frame
gun_deaths_2 <- gun_deaths %>%
    mutate(season = case_when( (month %in% c("Jan", "Feb", "Mar")) ~ "Winter",
                               (month %in% c("Apr", "May", "Jun")) ~ "Spring",
                               (month %in% c("Jul", "Aug", "Sep")) ~ "Summer",
                               (month %in% c("Oct", "Nov", "Dec")) ~ "Fall"))

# print out the season with most gun deaths
gun_deaths_2 %>%
    group_by(season) %>%
    count() %>%
    arrange(desc(n)) %>%
    ungroup() %>%
    top_n(1, n)
```
** Summer is the season with most most gun deaths

# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
gun_deaths_3 <- gun_deaths %>%
    filter(intent %in% c("Suicide", "Homicide") & race %in% c("White", "Black", "Hispanic")) 

ggplot(gun_deaths_3, aes(x = race, fill = intent)) + 
    geom_bar(position = "dodge") + 
    xlab("\nRace") +
    ylab("Number of Death\n") + 
    scale_y_continuous(breaks=seq(0, 60000, by = 10000)) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

(hint maybe looking at percentages would be good)
```{r race_2}
# summarize percentage of intent within each race group, limit race groups to White, Black, and Hispanic
gun_deaths %>%
    group_by(race, intent) %>%
    count() %>%
    group_by(race) %>%
    mutate(percent = n / sum(n) ) %>%
    select(race, intent, percent) %>%
    filter(intent %in% c("Suicide", "Homicide") & race %in% c("White", "Black", "Hispanic")) 
```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}
# convert police-involved from numeric to factor
gun_deaths$police <- factor(as.character(gun_deaths$police))
summary(gun_deaths$police)

# select variables needed for analysis and remove missng values
police <- gun_deaths %>%
    select(police, age, race) %>%
    filter(is.na(age) == FALSE) 

# calculate average age by police-involved
avg_age_police <- police %>%
    group_by(police) %>%
    summarize(mean_age=mean(age, na.rm = TRUE)) 

# round to the nearest tenth
avg_age_police$mean_age <- round(avg_age_police$mean_age, digits = 2)

# draw box plot of age by police-involved
ggplot(police, aes(police, age)) +
    geom_boxplot(color="blue") +
    xlab("\nPolice-involved") + 
    ylab("Age\n") +
    scale_y_continuous(breaks=seq(0, 100, by = 10)) + 
    stat_summary(fun = mean, colour = "red", geom = "point", shape = 18, size = 2) +
    geom_text_repel(avg_age_police, mapping = aes(x = police, y = mean_age, label = mean_age), inherit.aes = FALSE) + 
    scale_x_discrete(breaks=c("0", "1"), labels=c("Uninvolved", "Involved"))
```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}
# use t-test to check if the differences between two means are statistically significant
t.test(police$age ~ police$police)
```

**The results of t-test suggest that the differences are statistically significant

```{r police_2}
# create a new data set with percentage of police-involved death within each race group
police_2 <- police %>%
    group_by(race, police) %>%
    count() %>%
    ungroup() %>%
    group_by(race) %>%
    mutate(percent = n / sum(n) * 100) %>%
    ungroup()

# round percent to the tenth
police_2$percent = round(police_2$percent, digit = 2)

# draw a bar chart to visualize the differences across race groups
ggplot(police_2, aes(race, percent, fill = police)) + 
    geom_col(position = "dodge") + 
    xlab("\nRace") +
    ylab("Percent of Police-involved Death\n") + 
    scale_y_continuous(breaks=seq(0, 100, by = 10)) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r stats_2?}
# use ANOVA to check if the differences across groups are statistically significant
police$police <- as.numeric(police$police)
res.aov <- aov((police ~ race), data = police)
summary(res.aov)
```


Finally, all together now:
```{r police_3}
#convert police to factor
gun_deaths$police <- factor(as.character(gun_deaths$police))
summary(gun_deaths$police)

# calculate average age by police-involved
avg_age_police <- police %>%
    group_by(race, police) %>%
    summarize(mean_age=mean(age)) 

# round to the nearest tenth
avg_age_police$mean_age <- round(avg_age_police$mean_age, digits = 2)

ggplot(police, aes(factor(police), age)) +
    geom_boxplot(color = "orange") +
    facet_wrap(~race, nrow = 1, labeller = label_wrap_gen(width=18)) + 
    xlab("\nPolice-involved") + 
    ylab("Age\n") +
    scale_y_continuous(breaks=seq(0, 100, by = 10)) + 
    stat_summary(fun = mean, colour = "red", geom = "point", shape = 18, size = 2) +
    geom_text_repel(avg_age_police, mapping = aes(x = police, y = mean_age, label = mean_age), inherit.aes = FALSE) + 
    scale_x_discrete(breaks=c("0", "1"), labels=c("Uninvolved", "Involved"))
```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

