---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(knitr)
library(dplyr)
library(ggplot2)

# read in the data
guns <- read.csv(file = "C://Users//danie//Research Data//Git//HW04-2/gun_deaths.csv", header = TRUE)

view(guns) ##to print out data frame and look at how each column is classified
```
## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
    #month_chart <- guns %>% 
    #group_by(month) %>%
    #summarise(total = sum(month))

##I don't know why I get this comment that 'summarise()' ungrouping output (override with 'groups' argument ... if I swich the two functions around, I get an error. If I omit group_by and only run the summarise function, I get a total of all the deaths combined (at least only one datum output .. the code is running, so I'm leaving it).

## REVISITING, I SEEMED TO HAVE SUMMED UP MONTH AFTER MONTH, NOT A SUM FOR EACH INVIDIUAL MONTH (meaning December basically had the total for the deaths combine). ; then deleted basic knitr function. Took a peak at someone else's homework assignment, and tried to follow along as well as with a Google Answer I saw using this format. 

knitr::kable(guns %>% count(month), format = "markdown")
#finally

```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
ggplot(guns, aes(x= month)) + geom_bar() + 
##this yields a graph that has the columns on a scale (so no true 1, 2, 3, 4.. etc) Instead the default x-axis was 0-12.5   

scale_x_discrete(labels = c("January", "February" , "March" , "April" , "May" , "June" , "July" , "August" , "September" , "October" , "November", "December"))
##this didn't solve the problem, but I'm tired. 
   


##BELOW IS GARBAGE AKA MY FIRST ATTEMPT AT THIS 
    #monthly <- table(month_chart$month)
    #monthly
    #barplot(monthly)

##this barplot() was giving the bars labeled 1, 2, 3 etc on the x-axis, but the y-axis is now out of scale and I am unsure how to scale it without further Googling. 

##after printing out monthly, each month has an additional value of 1 for some reason. I tried to convert the dataframe into a matrix. but now it gives only two bars. 

    #monthly <- data.matrix(month_chart)
    #barplot(month_only)

```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
intent <- guns %>%
    count(intent)

intent

ggplot(intent, aes(x = reorder(intent, -n) , y = n)) + geom_col() +labs(x = "Intent of Gun", y = "Total Number of Victims") #labs function added 07/30/2020 
##reorder trick picked up from Google search
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
guns$age <- as.numeric(guns$age) #I stole this from a classmate's work. I saw that without this, my boxplot was giving a value to each age and skewing the overall look of the boxplot, as well being illegible. 

ggplot(guns, aes(x= sex,  y = age)) + geom_boxplot() + labs(x = "Sex of Victim" , y ="Age", title = "Boxplot of Sex vs Age of Gun Vicitims from 2012 -2014") #labs function added 07/30/2020 

all_age <- guns %>%
    select(sex, age) %>%
    filter(sex == "F")

all_age <- all_age[complete.cases(all_age), ]  #I remembered this trick from my assignment last week. Withouth this statement, taking the mean of the Female ages was yielding an NA result. Presumably some values are not provided (no age given) so I used complete cases to only include those that had a value. After I wrote this in, mean function returned a number. 

mean(all_age$age)
```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
education <- guns %>% 
    filter(sex == "M") %>%
    filter(year == 2012) %>%
    filter(education != "Less than HS") %>%
    filter(race == "White")

nrow(education)

#the answer is 1548 white males with at least a high school education were killed by guns in 2012
```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
Winter <- guns %>% filter(month==c(1:3))
Spring <- guns %>% filter(month==c(4:6))
Summer <- guns %>% filter(month==c(7:9))
Fall <- guns %>% filter(month==c(10:12))

nrow(Winter) #=7824
nrow(Spring) #=8606
nrow(Summer) #=8772
nrow(Fall) #=8401

#Summer has the most deaths
##I submitted this code on 07/30/2020 , may be the lazy way of (since I am not visually explaining the answer. But I used nrows to see how many rows were in each of the data frames I've made)
```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
white_race <- guns %>% filter(race == "White")
white_intent <- white_race %>% count(intent)
ggplot(white_intent, aes(x=reorder(intent, -n), y=n)) + geom_col() + labs(x="Intent of Gun" , y = "Numbers of Individuals" , title = "Bar Graph of White Individuals and Intent of Gun Deaths")

#labs function added 07/30/2020
```

(hint maybe looking at percentages would be good)
```{r race_2}
black_race <- guns %>% filter(race == "Black")
black_intent <- black_race %>% count(intent)
ggplot(black_intent, aes(x=reorder(intent, -n), y=n)) + geom_col() + labs(x="Intent of Gun" , y = "Numbers of Individuals" , title = "Bar Graph of Black Individuals and Intent of Gun Deaths")

hispanic_race <- guns %>% filter(race == "Hispanic")
hispanic_intent <- hispanic_race %>% count(intent)
ggplot(hispanic_intent, aes(x=reorder(intent, -n), y=n)) + geom_col() + labs(x="Intent of Gun" , y = "Numbers of Individuals" , title = "Bar Graph of Hispanic Individuals and Intent of Gun Deaths")


#more gun victims who were white persons died because of suicide, while black and hispanic persons died because of homicide

```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}
guns$age <- as.numeric(guns$age)

black_stat<- guns %>% filter (race == "Black")
white_stat<- guns %>% filter (race == "White")


ggplot(black_stat, aes(x= police)) + geom_bar()

ggplot(white_stat, aes(x= police)) + geom_bar()
```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}


```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

