---
title: "Exploring Gun Deaths in America"
author: Patrick Cantwell
date: "7/27/2020"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library("knitr")
library("devtools")

# read in the data
gun_deaths <- read.csv("gun_deaths.csv") # I decided to store this as a data frame
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}

month_abbrev <-as.factor(month.abb) #creates a factor class vector of the built-in character vector containing abbrevations of the months

gun_deaths_months_kable <- gun_deaths %>% 
    select(year, month) %>% 
    group_by(month) %>% 
    count(month)

gun_deaths_months_kable$month <- month_abbrev #changes the numbers that represent the months into their abbreviations

kable(gun_deaths_months_kable, format = "markdown", col.names = c("Month", "# of Deaths"), caption = "Sum of deaths for each month from 2012-2014")

```




### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
gun_deaths_months_kable$month = factor(gun_deaths_months_kable$month, levels = month.abb) # to set the order of the factor
gun_deaths_months_kable %>% 
    ggplot(aes(x = month, y = n)) +
    geom_col() +
    labs(title = "Sum of deaths for each month from 2012-2014",
         x = "Month",
         y = "Number of Deaths")
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
gun_deaths_intent <- gun_deaths %>% 
    select(intent) %>% 
    count(intent) %>% 
    arrange(desc(n))

gun_deaths_intent$intent = factor(gun_deaths_intent$intent)

gun_deaths_intent %>% 
    ggplot(aes(x = reorder(intent, -n), y = n)) + #using the reorder() function with aes() allows for the reordering of the x-axis based on y-value
    geom_col() +
    labs(title = "Gun Deaths by Intent",
         x = "Intent",
         y = "# of Deaths")
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
gun_death_age_sex <- gun_deaths %>% 
    select(age, sex) %>% 
    group_by(sex, age) %>% 
    filter(age != "Not Available") %>% 
    type.convert() #This break in the code forces the age column to be numeric
    
gun_death_age_sex2 <- gun_death_age_sex %>% 
    count(age) %>% 
    arrange(age)

gun_death_age_sex2 %>% 
    ggplot(aes(x = sex, y = age)) +
    geom_boxplot() +
    labs(title = "Age at Death of Gun Victims by Sex, 2012-2014",
         x = "Sex",
         y = "Age")
    
```

This code shows that the average age of death for female victims of gun violence is ~43.7 years old. 
```{r}
gun_death_age_sex3 <- gun_death_age_sex %>% 
    filter(sex == "F") %>% 
    ungroup(sex, age) %>% 
    summarize(average_age_F = mean(age))

print(gun_death_age_sex3$average_age_F[1])
```



I kind of feel like I didn't do this last graph correctly, so I decided to graph this data in another way as well. 
```{r age_&_sex_alternate}
gun_death_age_sex2 %>% 
    ggplot(aes(x = age, y = n, fill = sex)) +
    geom_col() +
    facet_grid(rows = vars(sex)) +
    labs(title = "Gun Deaths by Age and Sex, 2012-2014",
         x = "Age",
         y = "# of Deaths") +
    scale_y_continuous(expand = c(0,0), limits = c(0, 2750)) +
    scale_x_continuous(expand = c(0,0), limits = c(0, 100)) #while there are victims over the age of 100, they would not appear on the graph anyway
```



## How many white males with at least a high school education were killed by guns in 2012?

```{r education, warning= FALSE} 
gun_deaths_e_r_s <- gun_deaths %>% 
    select(year, education, race, sex) %>% 
    filter(education == c("BA+", "Some college", "HS/GED"), race == "White", sex == "M", year == "2012") %>% 
    count()

print(gun_deaths_e_r_s$n[1])
    
```
**5,078** white males with at least a high school education were killed by guns in 2012.

## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
gun_deaths_season <- gun_deaths %>% 
    select(month) %>% 
    count(month)
    
gun_deaths_season$month <- month_abbrev

gun_deaths_season2 <- gun_deaths_season %>% 
    summarize(Winter = sum(n[1:3]), Spring = sum(n[4:6]), Summer = sum(n[7:9]), Fall = sum(n[10:12]))

kable(gun_deaths_season2, format = "markdown")
```

This table shows that Summer has the most total gun deaths from 2012-2014

# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
gun_death_s_h <- gun_deaths %>% 
    select(race, intent) %>% 
    filter(race == c("White", "Black", "Hispanic"), intent == c("Suicide", "Homicide")) %>% 
    group_by(race) %>% 
    count(intent) %>% 
    mutate(total_death = sum(n)) %>% 
    mutate(fraction_intent = n / total_death)

gun_death_s_h %>% 
    ggplot(aes(x = race, y = fraction_intent, fill = intent)) +
    geom_col() +
    labs(title = "proportion of Gun Deaths that are Homicides or Suicides by Race, 2012-2014",
         x = "Race",
         y = "Proportion") +
    scale_y_continuous(expand = c(0,0))
```

The graph shows that a White person killed by a gun is more likely to die by suicide, while a Black or Hispanic person who is killed by a gun is more likely to die by homicide.


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}
police_factor <- c("Uninvolved", "Involved")
gd_police_race <- gun_deaths %>% 
    select(age, race, police) %>% 
    filter(age != "Not Available") %>% 
    group_by(race) %>% 
    count(police) %>% 
    mutate(total_death = sum(n)) %>% 
    mutate(fraction = n / total_death)

gd_police_race$police <- as.factor(rep.int(police_factor, 5)) #Change police column to factor

gd_police_race2 <- gd_police_race %>% 
    filter(police == "Involved")


gd_police_race2 %>%  #Definitely not pretty, but it will do
    ggplot(aes(x = reorder(race, fraction), y = fraction, fill = "red4")) +
    geom_col() +
    labs(title = "Fraction of Gun Death that Involve Police by Race",
        y = "Fraction of Gun Deaths Involving Police",
        x = "Race of Victim") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1)) + #There are packages that allow for the axis text to wrap, but I decided not to use it
    theme(legend.position = "none")

```


```{r police_2}
gd_police_age <- gun_deaths %>% 
    select(age, police) %>% 
    filter(age != "Not Available") %>% 
    mutate(age = as.numeric(age), police = as.factor(police)) %>% 
    group_by(age) %>% 
    count(police) %>% 
    mutate(total_death = sum(n)) %>% 
    mutate(fraction = n / total_death)

levels(gd_police_age$police) <- police_factor

gd_police_age2 <- gd_police_age %>% 
    filter(police == "Involved")

gd_police_age2 %>% # missing columns represent zero police-involved deaths for that age
    ggplot(aes(x = age, y = fraction, fill = "red4")) +
    geom_col() +
    labs(title = "Fraction of Gun Death that Involve Police by Age",
        y = "Fraction of Gun Deaths Involving Police",
        x = "Age of Victim") +
    theme(legend.position = "none")

```



Finally, all together now:
```{r police_3}
gd_police_a_r <- gun_deaths %>% 
    select(age, race, police) %>% 
    filter(age != "Not Available", race == c("White", "Black", "Hispanic")) %>% #race was filted because 
    mutate(age = as.numeric(age), police = as.factor(police), race = as.factor(race)) %>% 
    group_by(race, age) %>% 
    count(police) %>% 
    mutate(total_death = sum(n)) %>% 
    mutate(fraction = n / total_death)

levels(gd_police_a_r$police) <- police_factor

gd_police_a_r2 <- gd_police_a_r %>% 
    filter(police == "Involved")

gd_police_a_r2 %>% 
    ggplot(aes(x = age, y = fraction, fill = race)) +
    geom_col()+
    facet_grid(rows = vars(race)) +
    theme(legend.position = "none")
```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

