---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2

# read in the data
gun <- read.csv("gun_deaths.csv", header = TRUE, stringsAsFactors = FALSE)
head(gun)

```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
library(knitr)
#year <- gun$year
#month <- gun$month
#year_month <- paste(year,month,sep="/")
#year_month <- table(year_month) 
#year_month <- as.data.frame(year_month)
#colnames(year_month) <- c("Year_month", "Gun_Death")
#kable(year_month, format = "markdown")

month_gun <- table(gun$month)
month_gun <- as.data.frame(month_gun)
colnames(month_gun) <- c("month","gun_death")
kable(month_gun, format = "markdown")

```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
month_gun$month <- month.name
#need to make month an ordered factor, so ggplot doesn't do it alphabaticially
#Factors are used to represent categorical data. Factors can be ordered or unordered and are an important class for statistical analysis and for plotting.
month_gun$month <- factor(month_gun$month, levels = month_gun$month)

ggplot(month_gun)+
    geom_bar(aes(x = month, y = gun_death, fill = month), stat = "identity")+
    labs(x = "Month", y = "Gun death per month", title = "Gun death per month")+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"), 
        legend.position = "none")

```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}

gun_intent <- gun$intent
gun_intent <- as.data.frame(table(gun_intent))
colnames(gun_intent) <- c("intent","death")

ggplot(gun_intent)+
    geom_bar(aes(x = intent,  y = death, fill = intent), stat = "identity")+
    labs(x = "Intent", y = "Gun death", title = "Gun death for each intent")+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"), 
        legend.position = "none")

```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}

gun_age_sex <- gun[,5:6]
#gun_age_sex <- as.data.frame(table(gun_age_sex))
#colnames(gun_age_sex) <- c("sex", "age", "death")

#remove the not available line
gun_age_sex <- subset(gun_age_sex, age != "Not Available")
#gun_age_sex <- subset(gun_age_sex, death != 0)
gun_age_sex$age <- as.numeric(gun_age_sex$age)

#draw the plot
ggplot(gun_age_sex)+
    geom_boxplot(aes(x = sex, y = age, fill = sex))

#calculate average female age
gun_age_F <- subset(gun_age_sex, sex == "F")
mean(gun_age_F$age)

```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}

gun_white_2012 <- subset(gun, race == "White" & year == "2012")
gun_white_2012 <- subset(gun_white_2012, education != "Not Available" & education != "Less than HS")
#check to make sure all less HS are filtered out
table(gun_white_2012$education)
#print out number of white males with at least a high school education were killed by guns in 2012
nrow(gun_white_2012)

```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
gun$month[gun$month == "1"] <- "Winter"
gun$month[gun$month == "2"] <- "Winter"
gun$month[gun$month == "3"] <- "Winter"
gun$month[gun$month == "4"] <- "Spring"
gun$month[gun$month == "5"] <- "Spring"
gun$month[gun$month == "6"] <- "Spring"
gun$month[gun$month == "7"] <- "Summer"
gun$month[gun$month == "8"] <- "Summer"
gun$month[gun$month == "9"] <- "Summer"
gun$month[gun$month == "10"] <- "Fall"
gun$month[gun$month == "11"] <- "Fall"
gun$month[gun$month == "12"] <- "Fall"
#is there faster way to do this?, instead of type out all of them?

summary <- as.data.frame(table(gun$month))
season_order <- order(summary$Freq, decreasing=TRUE)
gun$month[season_order[1]]

```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
#for white
white_gun <- subset(gun, race == "White")
white_gun <- white_gun["intent"]
white_gun <- as.data.frame(table(white_gun))
colnames(white_gun) <- c("intent", "death")

suicide <- which(white_gun$intent == "Suicide")
homicide <- which(white_gun$intent == "Homicide")

white_gun[suicide, 2] > white_gun[homicide, 2]


#for black
black_gun <- subset(gun, race == "Black")
black_gun <- black_gun["intent"]
black_gun <- as.data.frame(table(black_gun))
colnames(black_gun) <- c("intent", "death")

suicide <- which(black_gun$intent == "Suicide")
homicide <- which(black_gun$intent == "Homicide")

black_gun[suicide, 2] > black_gun[homicide, 2]


#for hispanics
hispanic_gun <- subset(gun, race == "Hispanic")
hispanic_gun <- hispanic_gun["intent"]
hispanic_gun <- as.data.frame(table(hispanic_gun))
colnames(hispanic_gun) <- c("intent", "death")

suicide <- which(hispanic_gun$intent == "Suicide")
homicide <- which(hispanic_gun$intent == "Homicide")

hispanic_gun[suicide, 2] > hispanic_gun[homicide, 2]

```

(hint maybe looking at percentages would be good)
```{r race_2}

```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}
gun_police_age_race <- gun[,c("police", "race", "age")]
gun_police_age_race <- subset(gun_police_age_race, age != "Not Available")


#police involvement and age
gun_police_age <- gun_police_age_race[,c("police", "age")]
gun_police_age_1 <- subset(gun_police_age, police == 1)
#mean age with police involvement
mean(as.numeric(gun_police_age_1[,"age"]))

gun_police_age_0 <- subset(gun_police_age, police == 0)
#mean age without police involvement
mean(as.numeric(gun_police_age_0[,"age"]))



#police involvement and race
gun_police_race <- gun_police_age_race[,c("police", "race")]
gun_police_race <- as.data.frame(table(gun_police_race))
#get percentage of police involvement for each race can solve the problem

which(gun_police_race$race == "Asian/Pacific Islander")
which(gun_police_race$race == "Black")
which(gun_police_race$race == "Hispanic")
which(gun_police_race$race == "Not Availabletive American/Not Availabletive Alaskan")
which(gun_police_race$race == "White")



```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}


```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

