---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(magrittr)    # For pipe sign
library(knitr)    # To pretty print tables
library(ggpubr)    # For p-value bars

# read in the data
gun_deaths = read.csv("gun_deaths.csv")

```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
deaths_per_month = gun_deaths %>%
    count(month)
kable(deaths_per_month, format = "markdown")

```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
month_labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sept", "Oct", "Nov", "Dec")
deaths_per_month %>%
    mutate(month_lab = month_labels[month]) %>%
    ggplot(aes(x = reorder(month_lab, month), y = n)) +
    geom_bar(stat = "identity")    # 'identity' means you've already done the summarization
    #scale_x_continuous(breaks = 1:12, labels = month_labels)

# The easier way do to exactly the same thing:
# Set 'month' as x aesthetics, and add the last commented line
# However, it is fun to achieve this by mutating the original tibble

```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
deaths_by_intent = gun_deaths %>%
    count(intent)
deaths_by_intent %>%    # Not sure why cannot pipe here
    ggplot(aes(x = reorder(intent, -n), y = n)) +    # -n means descending sort by n
    geom_bar(stat = "identity")

```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex, warning=FALSE}
gun_deaths_num_age = gun_deaths %>%
    # avoid unexpected behavoir when directly converting factors to numerics
    mutate(num_age = as.numeric(as.character(age))) %>%
    # also rm NA
    filter(is.na(num_age) == FALSE)
gun_deaths_num_age %>%
    ggplot(aes(x = sex, y = num_age)) +
    geom_boxplot()
gun_deaths_num_age %>%
    filter(sex == "F") %>%
    summarize(ave_age = mean(num_age)) %>%
    print()

```
Above is the average age of female gun death victims


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
gun_deaths %>%
    filter(race == "White", sex == "M",
           education %in% c("BA+", "HS/GED", "Some college"),
           year == 2012) %>%
    count() %>%
    print()

```
Above is the number of white... in 2012 got killed


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
season_labs = c("Winter", "Spring", "Summer", "Fall")
gun_deaths %>%
    mutate(season = season_labs[as.integer((month - 1) / 3) + 1]) %>%
    count(season) %>%
    arrange(desc(season)) %>%
    top_n(1, season) %>%
    print()
    
```
Above is the season with the most gun deaths.


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
intent_count = gun_deaths %>%
    filter(intent == "Homicide" | intent == "Suicide",
           # In C/C++, one may want to use || and && for logic OR and AND.
           # However in R, | is different from ||, that | can return a vector of Booleans. | is usually preferrred
           race %in% c("White", "Hispanic", "Black")) %>%
    group_by(race) %>%
    count(intent) 
intent_count %>%
    ggplot(aes(x = race, y = n, fill = intent)) +
    geom_bar(position = "dodge", stat = "identity")

```

(hint maybe looking at percentages would be good)
```{r race_2}
# Actually, I think the bar plot above is quite abvious
intent_count %>%
    # Notice it is still grouped by race
    transmute(race, intent, intent_percent = n / sum(n) * 100) %>%
    kable(format = "markdown")

```
Above shows the percent of homi- / sui- cide among different racial groups.


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police vs age}
# Age in involved/not involved groups
gun_deaths_num_age %>%
    ggplot(aes(x = as.factor(police), y = num_age)) +
    geom_boxplot() +
    scale_x_discrete(labels = c("not involved", "involved"), breaks = c(0, 1)) +
    labs(title = "Age in police involved/not involved groups")
```


```{r police vs race, warning=FALSE}
# Num of deaths in involved/not involved groups of different races
gun_deaths_modified = gun_deaths_num_age %>%
    mutate(police_lab = c("not involved", "involved")[police + 1]) %>%    # 0=not involved; 1=involved
    mutate(race = ifelse(race == "Not Availabletive American/Not Availabletive Alaskan",
                           "NA", as.character(race)))    # This one is too long for x-axis labels

gun_deaths_modified %>%
    group_by(police_lab, race) %>%
    summarize(race_total = n()) %>%
    # Still grouped by police
    mutate(race_percent = race_total / sum(race_total) * 100) %>%
    ggplot(aes(x = race, y = race_percent, fill = police_lab)) +
    geom_bar(position = "dodge", stat = "identity") +
    labs(title = "Race percentage in police involved/not involved groups") +
    # Tilt xtick labels to avoid overlap
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

```

```{r all three factors}
# This is the first idea that comes off the top of my head
# Stratifying the "age" column may give bar plots, should also be good
gun_deaths_modified %>%
    ggplot(aes(x = race, y = num_age, fill = police_lab)) +
    geom_boxplot(position = "dodge") +
    labs(title = "Age in police involved/not involved, per race")
```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}
t.test(gun_deaths_modified$num_age ~ gun_deaths_modified$police)


```
Above is the t-test telling whether there is a significant difference between the average age of the group "police involved" and "police not involved".

The conclusion is *significant* (p<0.05).


```{r police_2}
count_by_police_and_race = gun_deaths_modified %>%
    group_by(police_lab, race) %>%
    count() %>%
    # Now convert "race" to columns, i.e., make the table wider, using `spread()`
    spread(race, n)   # tibble format actually lost after calling `spread`. This is strange since `spread` is a tidyr func
kable(count_by_police_and_race, format = "markdown")
chisq.test(as.matrix(count_by_police_and_race[-1]))    # remove the "row name" column

```
Above Chi-square test inquires whether the number of gun deaths in different racial groups is independent of police involvement.

I just left one R coding block blank here. Similar to the previous t-test, we are able to perform t-test in groups of different police involvement of various races.


Finally, all together now:

Add the p-value of the t-test directly on the graph.
```{r police_3}
gun_deaths_num_age %>%
    ggplot(aes(x = as.factor(police), y = num_age)) +
    geom_boxplot() +
    scale_x_discrete(labels = c("not involved", "involved"), breaks = c(0, 1)) +
    labs(title = "Age in police involved/not involved groups") +
    stat_compare_means(method = "t.test")

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

