---
title: "Exploring Gun Deaths in America"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(viridis)      # color scheme
library(scales)       # required for overlaying percentages over bar plots

# set theme
theme_set(theme_bw())

my_colors <- c("#2a9d8f", "#E76F51", "#E9C46A", "#3a6b7e", "#F4A261")

# read in the data
gun_deaths <- read.csv('gun_deaths.csv', 
                       na.strings=c("NotAvailable"), 
                       colClasses = c("integer", "integer",  
                                      "factor", "factor",  
                                      "factor",  "integer",  
                                      "factor", "integer",  
                                      "factor",  "factor"))

# I thought the following would work, and I can't figure out why it doesn't.
# gun_deaths <- gun_deaths %>% mutate(race = replace(race, race == "Not Availabletive American/Not Availabletive Alaskan", "Native AmericaNative"))
# It just replaces my string with NA, not the new string. :(

# So I did this instead:
gun_deaths <- gun_deaths %>% mutate(race = gsub("Not Availabletive American/Not Availabletive Alaskan", "Native American/Native Alaskan", race))
gun_deaths <- gun_deaths %>% mutate(race = gsub("/", "/\n", race))

```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

kable is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
deaths_per_month <- gun_deaths %>% count(month)

knitr::kable(deaths_per_month, format = "markdown")
```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
ggplot(deaths_per_month, aes(month.abb[month], n)) +
  geom_bar(stat = "identity") +
  xlab("Month") +
  ylab("Gun Deaths") +
  scale_x_discrete(limits = month.abb)
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
deaths_per_intent <- gun_deaths %>% count(intent)

ggplot(deaths_per_intent, aes(reorder(intent, -n), n)) +
  geom_bar(stat = "identity") +
  xlab("Intent") +
  ylab("Gun Deaths")
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
ggplot(gun_deaths, aes(sex, age)) +
  geom_boxplot()
```

```{r female avg age}
avg_age_f <- gun_deaths %>% 
  group_by(sex) %>% 
  summarize(average_age = mean(age, na.rm=TRUE)) %>%
  filter(sex == "F")

paste("Average age of female gun death victims:", avg_age_f[[2]])
```

## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
count_m_white_hs <- gun_deaths %>%
  filter(race == "White" & 
           sex == "M" & 
           education != "Less than HS" & 
           year == 2012) %>%
  count()

paste("Number of white males with at least a HS education killed by guns in 2012:", 
      count_m_white_hs[[1]])
```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
gun_deaths_top_season <- fct_collapse(month.abb[gun_deaths$month],
  Winter = c("Jan", "Feb", "Mar"),
  Spring = c("Apr", "May", "Jun"),
  Summer = c("Jul", "Aug", "Sep"),
  Fall = c("Oct", "Nov", "Dec")) %>%
  fct_count() %>%
  top_n(1, n)

paste("Season with the most gun deaths:", gun_deaths_top_season[[1]])
```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
# Would nont have been able to figure out how to put percentages on the graphs without this tutorial:
# https://rkabacoff.github.io/datavis/Bivariate.html

gun_deaths_by_intent_and_race <- gun_deaths %>%
  group_by(race, intent) %>%
  summarize(n = n()) %>% 
  mutate(pct = n/sum(n),
         lbl = scales::percent(pct))
gun_deaths_by_intent_and_race

ggplot(gun_deaths_by_intent_and_race, aes(race, y = pct, fill = intent)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = lbl), size = 3, position = position_stack(vjust = 0.5)) +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=0.5)) +
  scale_y_continuous(breaks = seq(0, 1, .2), label = percent) +
  scale_fill_brewer(palette = "Set3", direction=1) +
  xlab("Race") +
  ylab("Percent of Deaths") +
  labs(fill = "Intent")
```

(hint maybe looking at percentages would be good)

Answer: Whites who are killed by guns are more likely to die of suicide than homicide. Black and hispanic people killed by guns are more likely to die of homicide.

### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police and age}
ggplot(gun_deaths, aes(factor(police, levels = c(0, 1), labels = c("No", "Yes")), age)) +
  geom_boxplot() +
  xlab("Police Involvement") +
  ylab("Age") +
  coord_flip()
```

```{r police and race}
gun_deaths_by_police_and_race <- gun_deaths %>%
  group_by(race, police) %>%
  summarize(n = n()) %>% 
  mutate(pct = n/sum(n),
         lbl = scales::percent(pct))

ggplot(gun_deaths_by_police_and_race, 
       aes(race, y = pct, fill = factor(police, levels = c(0, 1), labels = c("No", "Yes")))) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = lbl), size = 3, position = position_stack(vjust = 0.5)) +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=0.5)) +
  scale_y_continuous(breaks = seq(0, 1, .2), label = percent) +
  scale_fill_brewer(palette = "Set2") +
  xlab("Race") +
  ylab("Percent of Deaths") +
  labs(fill = "Police\nInvolvement")
```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

### Is police involvement correlated with age?

```{r stats police and age}
# Is age (independent variable) correlated with police involvement (dependent variable)?
# Test with logistic regression

mylogit <- glm(police ~ age, data = gun_deaths, family = "binomial")
summary(mylogit)

# p-value <2e-16 ***
# For every unit increase in age, the log odds of police involvement decrease by 0.02349.
```
Age has a significant negative correlation with police involvement.

### Is police involvement correlated with race?

```{r stats police and race}
# Is race (independent variable) correlated with police involvement (dependent variable)?
# Test with two-way chi squared

chisq <- chisq.test(x = table(gun_deaths$police, gun_deaths$race))


chisq$p.value
chisq$residuals

# p-value <2e-16 ***
# Race is significantly correlated with police involvement.
# There is a positive association between all races and police involvement, except for whites. 
# There is a negative association between being white and police involvement.
```

Race has a significant correlation with police involvement. 

## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

