---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(devtools)
library(knitr)
library(kableExtra)

# read in the data
gun_deaths_data <- read.csv("gun_deaths.csv")

#glimpsing at the data, couldn't use function "glimpse" for some reason
head(gun_deaths_data)

tail(gun_deaths_data)

summary(gun_deaths_data)


#changing label of months to three letter codes from digits 1-12
gun_deaths_data$month <- factor(gun_deaths_data$month, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec" ))

#changing race label
gun_deaths_data$race <- gsub(pattern = "Not Availabletive American/Not Availabletive Alaskan", 
     replacement = "Native American/Native Alaskan",
     x = gun_deaths_data$race)

```



## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
gun_deaths_per_month <- gun_deaths_data %>%
    group_by(month) %>%
    count()

kable(gun_deaths_per_month, format = "markdown", col.names = c("Month", "# of Gun Deaths"), caption = "number of deaths by gun in each month from 2012-2014, sourced from FiveThirtyEight")
```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
ggplot(gun_deaths_data, aes(month)) +
           scale_y_continuous(breaks = seq(0, 10000, by = 1000)) +
           geom_bar(fill = "orchid4") +
               labs(x = "Month", y = "Number of Gun Deaths", title = "Total Gun Deaths per Month from 2012-2014", caption = "Sourced from FiveThirtyEight") +
               theme_light()
           
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
ggplot(gun_deaths_data, aes(fct_infreq(intent))) +
    geom_bar(fill = "mediumspringgreen") +
    labs(title = "Gun Deaths Associated with Intent Cause of Death from 2012-2014", x = "Intent Cause of Death", y = "Number of Gun Deaths", caption = "Sourced from FiveThirtyEight") +
    theme_light()
    

```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
gun_deaths_data_by_sex <- gun_deaths_data %>%
    filter(age != "Not Available") %>%
    mutate(age = as.numeric(age))


ggplot(gun_deaths_data_by_sex, aes(sex, age)) +
    geom_boxplot(fill = "slategray3") +
    labs(title = "Age of Gun Death Victims by Sex from 2012-2014", x = "Sex", y = "Age of Gun Death", caption = "Sourced from FiveThirtyEight") +
    theme_light()

#Average age of female gun death victims printed below
gun_deaths_data_by_sex %>%
    filter(sex == "F") %>%
    summarize(avg_F_gundeath_age = mean(age))
    
```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
gun_deaths_data %>%
    filter(sex == "M", race == "White", education %in% c("BA+", "HS/GED", "Some college") , year == 2012) %>%
    count()
    
```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
gun_deaths_by_season <- gun_deaths_data %>%
    mutate(______________________ c("Jan", "Feb", "Mar") <- Winter, c("Apr", "May", "Jun") <- Spring, c("Jul", "Aug", "Sep") <- Summer, c("Oct", "Nov", "Dec") <- Fall ) %>%

count()

# not sure what to do here please help :')
```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}

```

(hint maybe looking at percentages would be good)
```{r race_2}

```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}

```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}


```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

