---
title: "Exploring Gun Deaths in America (2012 - 2014)"
output: github_document
---

## 1. Load the packages and read in the gun deaths data

```{r load-data, message=FALSE, warning=FALSE}

# Install packages

install.packages("ggsignif")

# Load the packages

library(tidyverse) 
library(dplyr)
library(knitr)
library(ggplot2)
library(data.table)
library(RColorBrewer)
library(ggsignif)

# Read in the data - fread() requires loading data.table package

gun_deaths <- fread("gun_deaths.csv")

```


## 2. Generate a data frame that summarizes the number of gun deaths per month in years 2012-2014 and print with kable()

```{r deaths per month - table, warning=FALSE}

# Sort the data by month

arranged_by_month <- gun_deaths %>%
    arrange(month)

# Change the numbers of the months to the abbreviations of months' names

arranged_by_month$month <- factor(arranged_by_month$month, c(1:12), labels = month.abb)

# Generate a table summarizing numbers of deaths in each month of the year

deaths_per_month <- table(arranged_by_month$month)

# Convert the table to a data frame

deaths_per_month_df <- data.frame(deaths_per_month)

# Print the table with modified column names; align the values in the columns

kable(deaths_per_month_df, col.names = c("Month", "Number of deaths"), align = "lc", format = "markdown")

```


## 3. Generate a bar chart showing the total number of gun deaths per month in years 2012-2014 

```{r deaths per month}

# Change the column names in the data frame

colnames(deaths_per_month_df) <- c("Month", "Number")

# Create a bar plot

bar_plot <- ggplot(data = deaths_per_month_df, aes(x = Month, y = Number, width = 0.75)) +
  
    geom_bar(stat = "identity", color = "black", fill = brewer.pal(length(unique(deaths_per_month_df$Month)), "Set3")) +
    
    labs(title = "Number of gun deaths in US by month (total for years 2012-14)", x = "Month", y = "Number of deaths") +

    theme(plot.title = element_text(size = 14, family = "Tahoma", face = "bold", hjust = 0.5),
          axis.title.x = element_text(colour="black", size = 12, face = "bold"),
          axis.title.y = element_text(colour="black", size = 12, face = "bold"),
          axis.text.x = element_text(colour="black", size = 10),
          axis.text.y = element_text(colour="black", size = 10),
          panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 8000, 9000, 10000), limits = c(0, 10000)) 

bar_plot

```

## 4. Generate a bar chart that identifies the number of gun deaths associated with each type of intent cause of death

```{r deaths by intent}

# Sort the data by intent

arranged_by_intent <- gun_deaths %>%
    arrange(intent)

# Generate a table summarizing deaths by intent

deaths_by_intent <- table(arranged_by_intent$intent)

# Convert the table to a data frame

deaths_by_intent_df <- data.frame(deaths_by_intent)

# Print the table with modified column names; align the values in the columns

kable(deaths_by_intent_df, col.names = c("Intent", "Number of gun deaths"), align = "lc", format = "markdown")

# Change the column names in the data frame

colnames(deaths_by_intent_df) <- c("Intent", "Number")

# Create a bar plot (specify the order of bars - from highest to lowest number of deaths)

bar_plot_intent <- ggplot(data = deaths_by_intent_df, aes(x = reorder(Intent, -Number), y = Number, width = 0.75, fill = Intent)) +
    
    geom_bar(stat = "identity", color = "black", fill = brewer.pal(length(unique(deaths_by_intent_df$Intent)), "Set3")) +
    
    labs(title = "Number of gun deaths in US by intent (total for years 2012-14)", x = "Intent", y = "Number of deaths") +

    theme(plot.title = element_text(size = 14, family = "Tahoma", face = "bold", hjust = 0.5),
          legend.position = "none", 
          axis.title.x = element_text(colour="black", size = 12, face = "bold"),
          axis.title.y = element_text(colour="black", size = 12, face = "bold"),
          axis.text.x = element_text(colour="black", size = 10),
          axis.text.y = element_text(colour="black", size = 10),
          panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 10000, 20000, 30000, 40000, 50000, 60000, 70000), limits = c(0, 70000)) 

bar_plot_intent

```

## 5. Generate a boxplot visualizing the age of gun death victims by sex

```{r deaths by age and sex}

# Generate a data frame with data about sex and age of victims; does not include the rows where the age is not available

by_sex_age <- gun_deaths %>%
    select(sex,age) %>%
    arrange(age) %>%
    filter(age != "Not Available") 

# Change age variable from factor to double

by_sex_age$age <- as.numeric(as.character(by_sex_age$age))

# Change the values in the sex column to include full words Female and Male

by_sex_age$sex <- factor(by_sex_age$sex, c("F", "M"), c("Female", "Male"))

# Generate a boxplot

ggplot(by_sex_age, aes(x = sex, y = age, fill = sex)) +
  
    geom_boxplot(outlier.shape = NA) +
    
    scale_fill_manual(values=c("#E69F00", "#56B4E9")) +
  
    labs(title = "Age of gun death victims by sex (for years 2012-14)", x = "Sex", y = "Age") +

    theme(plot.title = element_text(size = 14, family = "Tahoma", face = "bold", hjust = 0.5),
          legend.position = "none",
          axis.title.x = element_text(colour="black", size = 12, face = "bold"),
          axis.title.y = element_text(colour="black", size = 12, face = "bold"),
          axis.text.x = element_text(colour="black", size = 11, face = "bold"),
          axis.text.y = element_text(colour="black", size = 10),
          panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), limits = c(0, 100)) +

    annotate("text", x = 0.55, y = 43.7, label = "43.7", fontface = 2)

# Determine the average age of death for female victims

average_age_F <- by_sex_age %>%
    filter(sex == "Female") %>%
    summarize(women_average_age = mean(age))

average_age_F <- round(average_age_F,digits=2)

kable(average_age_F, col.names = c("Average age of female victims"), align = "lc", format = "markdown")

```

## 6. Determine the number of white males with at least a high school education who were killed by guns in 2012

```{r deaths by education}

# Create a data frame with data on the education of white males killed by guns in 2012 - include only high school and higher levels of education

white_M_high_school <- gun_deaths %>%
    select(year, sex, race, education) %>%
    filter(year == "2012", sex == "M", race == "White", education == c("BA+", "HS/GED", "Some college"))

# Determine the number of deaths

Number_of_M_HS_victims <- nrow(white_M_high_school)

kable(Number_of_M_HS_victims, col.names = c("Number of white male victims with at least a high school education killed by guns in 2012"), align = "lc", format = "markdown")

```
### _5063 white males with at least high school education were killed in 2012._

## 7. Determine which season of the year has the most gun deaths

```{r deaths by season}

# Create a summary of the number of deaths per month; use the abbreviations of the names of the months

deaths_months <- gun_deaths %>%
    select(month)

deaths_months$month <- factor(deaths_months$month, c(1:12), labels = month.abb)

# Print the table with data on number of deaths per month

kable(table(deaths_months), col.names = c("Month", "Number of deaths"), align = "lc", format = "markdown")

# Assign months to seasons

deaths_by_season <- deaths_months %>%
    mutate(season = month) %>%
    mutate(season = recode(season, 
                           Jan = "Winter",
                           Feb = "Winter",
                           Mar = "Winter",
                           Apr = "Spring",
                           May = "Spring",
                           Jun = "Spring",
                           Jul = "Summer",
                           Aug = "Summer",
                           Sep = "Summer",
                           Oct = "Fall",
                           Nov = "Fall",
                           Dec = "Fall")) %>%
    select(season)
    
# Create a data frame with a summary about the number of deaths per season; rename the columns

deaths_by_season_df <- data.frame(table(deaths_by_season))
colnames(deaths_by_season_df) <- c("Season", "Number")

# Print the table with data on number of deaths per season

kable(deaths_by_season_df, col.names = c("Season", "Number of deaths"), align = "lc", format = "markdown")

# Generate a bar plot

bar_plot_season <- ggplot(data = deaths_by_season_df, aes(x = Season, y = Number, width = 0.75, fill = Season)) +
  
    geom_bar(stat = "identity", color = "black", fill = brewer.pal(length(unique(deaths_by_season_df$Season)), "Set3")) +
  
    geom_text(aes(label = Number), position=position_dodge(width=1.5), vjust=-0.8) +
    
    labs(title = "Number of gun deaths in US by season (for years 2012-14)", x = "Season", y = "Number of deaths") +

    theme(plot.title = element_text(size = 14, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.position = "none", 
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 5000, 10000, 15000, 20000, 25000, 30000), limits = c(0, 30000)) 

bar_plot_season

```

### _Summer had the highest number of gun deaths across the years 2012-14._

## 8a. Determine whether Whites who are killed by guns are more likely to die because of suicide or homicide
##     Compare the result to Blacks and Hispanics

```{r deaths by race}

# Create a data frame with data on deaths by race and intent including only suicide and homicide

race_intent <- gun_deaths %>%
    select(race, intent) %>%
    filter(race == c("White", "Black", "Hispanic"), intent == c("Suicide", "Homicide"))

# Generate a data frame summarizing deaths by race and intent including suicide and homicide; rename the columns

race_intent_df <- data.frame(table(race_intent))

colnames(race_intent_df) <- c("Race", "Intent", "Number")

# Generate a bar plot

bar_plot_race_intent <- ggplot(data = race_intent_df, aes(x = Intent, y = Number, width = 0.9, fill = Race)) +
  
    geom_bar(stat = "identity", color = "black", position = "dodge2") +
  
    geom_text(aes(label = Number), position=position_dodge(width=0.9), vjust=-0.6, hjust = 0.5) +

    labs(title = "Number of males killed by guns because of suicide and homicide \nin US between 2012 and 2014 by race", x = "Intent", y = "Number of deaths") +

    theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 2000, 4000, 6000, 8000, 10000), limits = c(0, 10000)) 

bar_plot_race_intent

```

### _White people killed by guns were more likely to die due to suicide than homicide._
### _Black and Hispanic people killed by guns, however, died more often as a result of homicide than suicide._

## 8b. Determine what percentage of people killed by guns due to homicide and suicide correponds to Black, Hispanic, and White people. 

```{r deaths by race - percent}

# Calculate percentage of victims of different races killed due to homicide

race_intent_perc1 <- race_intent_df 

race_intent_perc1 <- race_intent_perc1 %>%
    filter(Intent == "Homicide") 

race_intent_perc1 <- race_intent_perc1 %>%
    mutate(Percentage = Number/sum(race_intent_perc1$Number))

# Calculate percentage of victims of different races killed due to suicide

race_intent_perc2 <- race_intent_df 

race_intent_perc2 <- race_intent_perc2 %>%
    filter(Intent == "Suicide")

race_intent_perc2 <- race_intent_perc2 %>%
    mutate(Percentage = Number/sum(race_intent_perc2$Number))

# Combine the data frames with calculated percentages

race_intent_perc3 <- rbind(race_intent_perc1, race_intent_perc2)   

# Generate bar plot

race_intent_perc_plot <- ggplot(race_intent_perc3, aes(x = Intent, y = Percentage, fill = Race)) +
 
   geom_bar(position = "fill", stat="identity", color = "black") +
  
   scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.00), labels = scales::percent) +

   labs(title = "Percentage of males killed by guns because of suicide and homicide \nin US between 2012 and 2014 by race", x = "Intent", y = "Number of deaths") +

   theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) 

race_intent_perc_plot

```

### _Among White, Black and Hispanic people killed by gun in 2012-2014 due to homicide, Black people comprised the highest percentage of victims (> 55%). Although White people comprised the second highest fraction of those victims, they were more likely to die due to suicide - they comprised around 90% of cases of suicide among these three races. Hispanics comprised around 17% of victims of homicide and around 5% of suicide cases._

### 9a. Assess the relationship between police involvement and age

```{r police involvement by age}

# Create a data frame and a summary with data on age of victims and police involvement

police_age <- gun_deaths %>%
    select(age, police)

police_age_df <- data.frame(table(police_age))

colnames(police_age_df) <- c("Age", "Police", "Number")

# Convert the values of age from factor to double

police_age_df$Age <- as.numeric(as.character(police_age_df$Age))

# Assign ranges of age to decades 

decade1 <- police_age_df %>%
    filter(Age %in% c(0: 10)) %>%
    mutate(Decade = "0-10")

decade2 <- police_age_df %>%
    filter(Age %in% c(11: 20)) %>%
    mutate(Decade = "11-20")

decade3 <- police_age_df %>%
    filter(Age %in% c(21: 30)) %>%
    mutate(Decade = "21-30")

decade4 <- police_age_df %>%
    filter(Age %in% c(31: 40)) %>%
    mutate(Decade = "31-40")

decade5 <- police_age_df %>%
    filter(Age %in% c(41: 50)) %>%
    mutate(Decade = "41-50")

decade6 <- police_age_df %>%
    filter(Age %in% c(51: 60)) %>%
    mutate(Decade = "51-60")

decade7 <- police_age_df %>%
    filter(Age %in% c(61: 70)) %>%
    mutate(Decade = "61-70")

decade8 <- police_age_df %>%
    filter(Age %in% c(71: 80)) %>%
    mutate(Decade = "71-80")

decade9 <- police_age_df %>%
    filter(Age %in% c(81: 90)) %>%
    mutate(Decade = "81-90")

decade10 <- police_age_df %>%
    filter(Age %in% c(91: 100)) %>%
    mutate(Decade = "91-100")

decade11 <- police_age_df %>%
    filter(Age %in% c(101: 110)) %>%
    mutate(Decade = "101-110")

by_decade <- rbind(decade1, decade2, decade3, decade4, decade5, decade6, decade7, decade8, decade9, decade10, decade11)

# Summarize the data on police involvement in death of victims of different ages

by_decade <- by_decade %>%
    select(Police, Number, Decade) %>%
    group_by(Decade, Police) %>%
    summarize(Number = sum(Number))

by_decade$Decade <- factor(by_decade$Decade,levels = c("0-10", "11-20",  "21-30", "31-40", "41-50", "51-60", "61-70",  "71-80", "81-90", "91-100", "101-110"))

# Plot number of deaths by age and police involvement

by_decade$Police <- factor(by_decade$Police, c("0", "1"), c("Not involved", "Involved"))

bar_plot_police_age1 <- ggplot(data = by_decade, aes(x = Decade, y = Number, width = 0.9, fill = Police)) +
  geom_bar(stat = "identity", color = "black", position = "dodge2") +
    
        labs(title = "Number of deaths by age and police involvement", x = "Age", y = "Number of deaths") +

    theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 4000, 8000, 12000, 16000, 20000, 24000), limits = c(0,24000)) 

bar_plot_police_age1

# Plot number of cases, in which police was not involved, by age

by_decade2 <- by_decade %>%
    filter(Police == "Not involved")

bar_plot_police_age2 <- ggplot(data = by_decade2, aes(x = Decade, y = Number, width = 0.6)) +
    geom_bar(stat = "identity", color = "black", position = "dodge2", fill = "indianred1") +
   
    labs(title = "Number of cases without police involvement by age", x = "Age", y = "Number of deaths") +

    theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 4000, 8000, 12000, 16000, 20000, 24000), limits = c(0,24000)) 

bar_plot_police_age2

# Plot number of cases, in which police was involved, by age

by_decade3 <- by_decade %>%
    filter(Police == "Involved")

bar_plot_police_age3 <- ggplot(data = by_decade3, aes(x = Decade, y = Number, width = 0.6)) +
    geom_bar(stat = "identity", color = "black", position = "dodge2", fill = "turquoise3") +
     
    labs(title = "Number of cases with police involvement by age", x = "Age", y = "Number of deaths") +

    theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 100, 200, 300, 400, 500, 600), limits = c(0,600)) 

bar_plot_police_age3

# Calculate the percentage of cases in which police was or was not involved by age

age_police_perc_all <- by_decade
  
age_police_perc_all <- age_police_perc_all %>%
    mutate(Percentage = Number/sum(age_police_perc_all$Number))

# Calculate the percentage of total cases, in which police was involved

police_involved <- age_police_perc_all %>%
  filter(Police == "Involved")

police_involved_perc <- sum(police_involved$Percentage)

kable(round(police_involved_perc, digits = 3), col.names = c("Fraction of cases, in which police was involved"), align = "l", format = "markdown")

# Plot the percentage of cases in which police was or was not involved by age

age_police_perc_plot4 <- ggplot(age_police_perc_all, aes(x = Decade, y = Percentage, fill = Police)) +
  geom_bar(position = "fill", stat="identity", color = "black") +
  scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.00), labels = scales::percent) +

 labs(title = "Percentage of cases in which police was and was not involved \nby age of the victim", x = "Age", y = "Percent of all deaths") +

    theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) 

age_police_perc_plot4

# Assess the difference between cases in which police was and was not involved in terms of age of the victims

age_police_perc5 <- by_decade %>%
    filter(Police == "Not involved")

age_police_perc5 <- age_police_perc5 %>%
    mutate(Percentage = Number/sum(age_police_perc5$Number))

age_police_perc6 <- by_decade %>%
    filter(Police == "Involved")

age_police_perc6 <- age_police_perc6 %>%
    mutate(Percentage = Number/sum(age_police_perc6$Number))

age_police_perc7 <- rbind(age_police_perc5, age_police_perc6)

age_police_perc_plot5 <- ggplot(age_police_perc7, aes(x = Police, y = Percentage, fill = Decade)) +
   
    geom_bar(position = "fill", stat="identity", color = "black") +
   
    scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.00), labels = scales::percent) +
    
    labs(title = "Percentage of cases in which police was and was not involved \nby age of the victim", x = "Police involvment", y = "Percent of all deaths") +

    theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) 

age_police_perc_plot5

```

### _Police was involved in less than 1.5% of total cases of death by guns. The absolute number of cases was the highest for people of age 21 to 30 independently from police involvement. Among the cases, in which police was involved, the highest percentage of victims were people of age 21 to 30. The second highest fraction were people of ages 31-40. Among the deaths, in which police was involved, the victims older than 60 comprised around than 5%, whereas among the cases, in which police was not involved, they comprised around 20%. Police was generally more involved in the cases of gun deaths among younger part of the population._

## Police involvement and race. 

```{r police involvement by race}

# Create a data frame with summary of police involvement in gun deaths of victims of different races

police_race <- gun_deaths %>%
    select(race, police)

by_race <- data.frame(table(police_race))
colnames(by_race) <- c("Race", "Police", "Number")

# Correct the name of "Native American/Native Alaskan"

by_race <- by_race %>%
    mutate(Race = recode(Race, 
                           "Not Availabletive American/Not Availabletive Alaskan" = "Native American/\nNative Alaskan",
                         "Asian/Pacific Islander" = "Asian/\nPacific Islander"))

race_police_perc1 <- by_race 
 
race_police_perc1 <- race_police_perc1 %>%
    mutate(Percentage = Number/sum(race_police_perc1$Number))

race_police_perc


race_police_perc_plot <- ggplot(race_police_perc, aes(x = Race, y = Percentage, fill = Police)) +
    geom_bar(position = "fill", stat="identity", color = "black") +
    scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1.00), labels = scales::percent) +

    labs(title = " race", x = "Intent", y = "Number of deaths") +

    theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) 

race_police_perc_plot

```

### 

## Police involvement by age and race. 

```{r police involvement by age and race}

#POLICE INVOLVED

police_race_age <- gun_deaths %>%
    select(age, race, police)

by_race_age_df <- data.frame(table(police_race_age))
colnames(by_race_age_df) <- c("Age", "Race", "Police", "Number")

by_race_age_df <- by_race_age_df %>%
    mutate(Race = recode(Race, 
                           "Not Availabletive American/Not Availabletive Alaskan" = "Native American/\nNative Alaskan",
                         "Asian/Pacific Islander" = "Asian/\nPacific Islander"))

by_race_age_df$Age <- as.numeric(as.character(by_race_age_df$Age))

decade1 <- by_race_age_df %>%
    filter(Age %in% c(0: 10)) %>%
    mutate(Decade = "0-10")

decade2 <- by_race_age_df %>%
    filter(Age %in% c(11: 20)) %>%
    mutate(Decade = "11-20")

decade3 <- by_race_age_df %>%
    filter(Age %in% c(21: 30)) %>%
    mutate(Decade = "21-30")

decade4 <- by_race_age_df %>%
    filter(Age %in% c(31: 40)) %>%
    mutate(Decade = "31-40")

decade5 <- by_race_age_df %>%
    filter(Age %in% c(41: 50)) %>%
    mutate(Decade = "41-50")

decade6 <- by_race_age_df %>%
    filter(Age %in% c(51: 60)) %>%
    mutate(Decade = "51-60")

decade7 <- by_race_age_df %>%
    filter(Age %in% c(61: 70)) %>%
    mutate(Decade = "61-70")

decade8 <- by_race_age_df %>%
    filter(Age %in% c(71: 80)) %>%
    mutate(Decade = "71-80")

decade9 <- by_race_age_df %>%
    filter(Age %in% c(81: 90)) %>%
    mutate(Decade = "81-90")

decade10 <- by_race_age_df %>%
    filter(Age %in% c(91: 100)) %>%
    mutate(Decade = "91-100")

decade11 <- by_race_age_df %>%
    filter(Age %in% c(101: 110)) %>%
    mutate(Decade = "101-110")

by_decade_age_race <- rbind(decade1, decade2, decade3, decade4, decade5, decade6, decade7, decade8, decade9, decade10, decade11)


by_decade_age_race <- by_decade_age_race %>%
    select(Police, Number, Decade, Race) %>%
    filter(Police == 1) %>%
    group_by(Decade, Race, Police) %>%
    summarize(Number = sum(Number))


by_decade_age_race


by_decade_age_race$Decade <- factor(by_decade_age_race$Decade,levels = c("0-10", "11-20",  "21-30", "31-40", "41-50", "51-60", "61-70",  "71-80", "81-90", "91-100", "101-110"))


bar_plot_police_age_race <- ggplot(data = by_decade_age_race, aes(x = Decade, y = Number, width = 0.9, fill = Race)) +
  geom_bar(stat = "identity", color = "black", position = "dodge2") +
    
        labs(title = "Police involved", x = "Age", y = "Number of deaths") +

    theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 20, 40, 60, 80, 100, 120, 140, 160, 180, 200), limits = c(0,200)) 


bar_plot_police_age_race


#Percent involved








#POLICE NOT INVOLVED


police_race_age <- gun_deaths %>%
    select(age, race, police)

by_race_age_df <- data.frame(table(police_race_age))
colnames(by_race_age_df) <- c("Age", "Race", "Police", "Number")

by_race_age_df <- by_race_age_df %>%
    mutate(Race = recode(Race, 
                           "Not Availabletive American/Not Availabletive Alaskan" = "Native American/\nNative Alaskan",
                         "Asian/Pacific Islander" = "Asian/\nPacific Islander"))

by_race_age_df$Age <- as.numeric(as.character(by_race_age_df$Age))

decade1 <- by_race_age_df %>%
    filter(Age %in% c(0: 10)) %>%
    mutate(Decade = "0-10")

decade2 <- by_race_age_df %>%
    filter(Age %in% c(11: 20)) %>%
    mutate(Decade = "11-20")

decade3 <- by_race_age_df %>%
    filter(Age %in% c(21: 30)) %>%
    mutate(Decade = "21-30")

decade4 <- by_race_age_df %>%
    filter(Age %in% c(31: 40)) %>%
    mutate(Decade = "31-40")

decade5 <- by_race_age_df %>%
    filter(Age %in% c(41: 50)) %>%
    mutate(Decade = "41-50")

decade6 <- by_race_age_df %>%
    filter(Age %in% c(51: 60)) %>%
    mutate(Decade = "51-60")

decade7 <- by_race_age_df %>%
    filter(Age %in% c(61: 70)) %>%
    mutate(Decade = "61-70")

decade8 <- by_race_age_df %>%
    filter(Age %in% c(71: 80)) %>%
    mutate(Decade = "71-80")

decade9 <- by_race_age_df %>%
    filter(Age %in% c(81: 90)) %>%
    mutate(Decade = "81-90")

decade10 <- by_race_age_df %>%
    filter(Age %in% c(91: 100)) %>%
    mutate(Decade = "91-100")

decade11 <- by_race_age_df %>%
    filter(Age %in% c(101: 110)) %>%
    mutate(Decade = "101-110")

by_decade_age_race <- rbind(decade1, decade2, decade3, decade4, decade5, decade6, decade7, decade8, decade9, decade10, decade11)


by_decade_age_race <- by_decade_age_race %>%
    select(Police, Number, Decade, Race) %>%
    filter(Police == 0) %>%
    group_by(Decade, Race, Police) %>%
    summarize(Number = sum(Number))


by_decade_age_race


by_decade_age_race$Decade <- factor(by_decade_age_race$Decade,levels = c("0-10", "11-20",  "21-30", "31-40", "41-50", "51-60", "61-70",  "71-80", "81-90", "91-100", "101-110"))


bar_plot_police_age_race <- ggplot(data = by_decade_age_race, aes(x = Decade, y = Number, width = 0.9, fill = Race)) +
  geom_bar(stat = "identity", color = "black", position = "dodge2") +
    
        labs(title = "Police not involved", x = "Age", y = "Number of deaths") +

    theme(plot.title = element_text(size = 13, family = "Tahoma", face = "bold", hjust = 0.5),
        legend.title = element_text(colour="black", size = 12, face = "bold"),
        axis.title.x = element_text(colour="black", size = 12, face = "bold"),
        axis.title.y = element_text(colour="black", size = 12, face = "bold"),
        axis.text.x = element_text(colour="black", size = 10, face = "bold"),
        axis.text.y = element_text(colour="black", size = 10),
        panel.grid.major.x = element_blank()) +
    
    scale_y_continuous(breaks = c(0, 2000, 4000, 6000, 8000, 10000, 12000, 14000), limits = c(0,14000)) 


bar_plot_police_age_race


#Percent - not involved











```

### _._

## Session info

```{r session}

devtools::session_info()

```

