---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library("knitr")

# read in the data
gun_deaths <- read.csv("gun_deaths.csv") # I decided to store this as a data frame
gun_deaths
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"

```{r}
str(gun_deaths)
summary(gun_deaths)
```


```{r month, warning=FALSE}

month_abbrev <-as.factor(month.abb) #creates a factor class vector of the built-in character vector containing abbrevations of the months

gun_deaths_months_kable <- gun_deaths %>% 
    select(year, month) %>% 
    group_by(month) %>% 
    count(month)

gun_deaths_months_kable$month <- month_abbrev #changes the numbers that represent the months into their abbreviations

kable(gun_deaths_months_kable, format = "markdown", col.names = c("Month", "# of Deaths"), caption = "Sum of deaths for each month from 2012-2014")

```




### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
gun_deaths_months_kable$month = factor(gun_deaths_months_kable$month, levels = month.abb) # to set the order of the factor
gun_deaths_months_kable %>% 
    ggplot(aes(x = month, y = n)) +
    geom_col() +
    labs(title = "Sum of deaths for each month from 2012-2014",
         x = "Month",
         y = "Number of Deaths")
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
gun_deaths_intent <- gun_deaths %>% 
    select(intent) %>% 
    count(intent) %>% 
    arrange(desc(n))

gun_deaths_intent$intent = factor(gun_deaths_intent$intent)

gun_deaths_intent %>% 
    ggplot(aes(x = reorder(intent, -n), y = n)) + #using the reorder() function with aes() allows for the reordering of the x-axis based on y-value
    geom_col() +
    labs(title = "Gun Deaths by Intent",
         x = "Intent",
         y = "# of Deaths")
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
gun_death_age_sex <- gun_deaths %>% 
    select(age, sex) %>% 
    group_by(sex, age) %>% 
    filter(age != "Not Available") %>% 
    type.convert() #This break in the code forces the age column to be numeric
    
gun_death_age_sex2 <- gun_death_age_sex %>% 
    count(age) %>% 
    arrange(age)

gun_death_age_sex2 %>% 
    ggplot(aes(x = sex, y = age)) +
    geom_boxplot() +
    labs(title = "Age at Death of Gun Victims by Sex, 2012-2014",
         x = "Sex",
         y = "Age")
    
```

This code shows that the average age of death for female victims of gun violence is ~43.7 years old. 
```{r}
gun_death_age_sex3 <- gun_death_age_sex %>% 
    filter(sex == "F") %>% 
    ungroup(sex, age) %>% 
    summarize(average_age_F = mean(age))

print(gun_death_age_sex3$average_age_F[1])
```



I kind of feel like I didn't do this last graph correctly, so I decided to graph this data in another way as well. 
```{r age_&_sex_alternate}
gun_death_age_sex2 %>% 
    ggplot(aes(x = age, y = n, fill = sex)) +
    geom_col() +
    facet_grid(rows = vars(sex)) +
    labs(title = "Gun Deaths by Age and Sex, 2012-2014",
         x = "Age",
         y = "# of Deaths") +
    scale_y_continuous(expand = c(0,0), limits = c(0, 2750)) +
    scale_x_continuous(expand = c(0,0), limits = c(0, 100)) #while there are victims over the age of 100, they would not appear on the graph anyway
```



## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
gun_deaths %>% 
    select(year, sex, race)
```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}

```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}

```

(hint maybe looking at percentages would be good)
```{r race_2}

```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}

```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}


```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

