---
title: "Exploring Gun Deaths in America"
output: github_document
---
Nabilah Sammudin
7/27/2020

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
# load tidyverse packages, including ggplot2
library(tidyverse) 
library(ggplot2)
library(knitr)
library (devtools)

#set the theme for all plots
theme_set(theme_classic())

# read in the data
gun_deaths <- read.csv("gun_deaths.csv")

# a glance at the data
glimpse(gun_deaths)
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
monthly_gun_deaths <- gun_deaths %>%
                        select(month) %>%
                        mutate(month = factor(month.abb[month],
                                              levels = c("Jan", "Feb", "Mar", "Apr",                                                         "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))) %>%
                        group_by(month) %>%
                        count()

kable(monthly_gun_deaths, align=rep('c'), format = "markdown",
      col.names = c("Month", "Total Deaths"),
      caption = "Number of Gun Deaths per Month")
```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
ggplot(monthly_gun_deaths, aes(month,n)) +
    geom_col(fill = "maroon") +
    labs(title = "Number of Gun Deaths per Month in the US (2012-2014)", 
       x = "Month", 
       y = "Number of Deaths") +
    theme(plot.title = element_text(hjust = 0.5))
    
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
gun_deaths %>%
    count(intent) %>%
    ggplot(mapping = aes(reorder(intent, -n), n)) +
    geom_col(fill = "maroon") +
    geom_text(aes(label = n), vjust = -0.5, size = 3, color = "grey50") +
    labs(title = "Number of Gun Deaths by Types of Intent in the US (2012-2014)", 
       x = "Types of Intent", 
       y = "Number of Deaths") +
    theme(plot.title = element_text(hjust = 0.5))
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
#remove NA age
sex_gun_deaths<- gun_deaths %>%
    filter(age != "Not Available") %>%
    mutate(age = as.numeric(age))

ggplot(sex_gun_deaths, mapping = aes(x = sex, y = age)) +
    geom_boxplot(fill = "maroon") +
     labs(title = "Ages of Gun Death Victims, By Sex",
          x = "Sex", 
          y = "Age (Years)") +
    scale_x_discrete(labels=c("Female", "Male")) +
    theme(plot.title = element_text(hjust = 0.5))

#determine average age of female gun death victims
sex_gun_deaths %>%
    filter(sex == "F") %>%
    summarize(average_female_age = mean(age))
```

**The average age of the female gun death victims in this dataset is 44.15 years.**
    
## How many white males with at least a high school education were killed by guns in 2012?

```{r education}

gun_deaths %>%
 filter(year == 2012, 
        race == "White", 
        education != "Less than HS",
        sex == "M") %>%
    count()
```
**15484 white males with at least a HS education were killed by guns in 2012.**

## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

**Answer**
Based on the previous table, summer has the most gun deaths.


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
filtered_race <- gun_deaths %>%
    filter(race %in% c("White", "Black", "Hispanic"), intent %in% c("Suicide", "Homicide")) %>%
    group_by(race, intent) %>%
    count() 

ggplot(filtered_race, mapping = aes(x = race, y = n)) +
           geom_col(aes(fill = intent)) +
           labs(title = "Distribution of Death Probability due to Suicide vs. Homicide by Races", x = "Races", y = "Number of Deaths") 
```
**Answer**
Based on this graph, whites who are killed by guns are more likely to die because of suicide. In other races (black and hispanic), they are more likely to get killed by guns due to homicide.

(hint maybe looking at percentages would be good)
```{r race_2}
filtered_race <- gun_deaths %>%
    filter(race %in% c("White", "Black", "Hispanic"), intent %in% c("Suicide", "Homicide")) %>%
    group_by(race, intent) %>%
    count(race) %>%
    group_by(race) %>%
    mutate(percentages = 100*n / sum(n))

ggplot(filtered_race, mapping = aes(x = race, y = percentages)) +
           geom_col(aes(fill = intent)) +
           labs(title = "Distribution of Death Probability due to Suicide vs. Homicide \nby Races (Percentages)", x = "Races", y = "Number of Deaths") +
        theme(plot.title = element_text(hjust = 0.5))
```
By looking at percentages based on this graph, ~ 85% whites who are killed by guns die because of suicide. Meanwhile, about ~ 85% blacks and ~ 70% Hispanics get killed by guns due to homicide.


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police and age}
police_gun_deaths <- gun_deaths %>%
    select(police, age, race) %>%
    filter(age != "Not Available") %>%
    mutate(age = as.numeric(age)) %>%
    group_by(police)

ggplot(police_gun_deaths, mapping = aes(x = factor(police), y = age)) +
    geom_boxplot(fill = "maroon") +
    labs(title = "Age Distribution of Gun Death Victims caused by Police Involvement ", 
          x = "Police Involvement", 
          y = "Age (Years)") +
    theme(plot.title = element_text(hjust = 0.5)) +
    scale_x_discrete(labels=c("Uninvolved", "Involved"))
```

```{r police and race}
new_police_gun_deaths <- gun_deaths %>%
    select(police, age, race) %>%
    filter(age != "Not Available") %>%
    mutate(age = as.numeric(age)) %>%
    group_by(police, race) %>%
    count(police) %>%
    group_by(police)

ggplot(new_police_gun_deaths, mapping = aes(x = factor(police), y = n)) +
    geom_col(aes(fill = race)) +
    labs(title = "Race Distribution of Gun Death Victims caused by Police Involvement ", 
          x = "Police Involvement", 
          y = "Race") +
    scale_x_discrete(labels=c("Uninvolved", "Involved"))

#This graph looks slightly off, I am not sure if the values for the 'involved' category is right. Will have to double-check!
```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}

age_police_uninvolved <- police_gun_deaths %>%
    filter(police == 0)
age_police_involved <- police_gun_deaths %>%
    filter(police == 1)

t.test(age_police_uninvolved$age, age_police_involved$age)
```
**Based on this t-test, the difference in age is statistically significant. 



Finally, all together now:
```{r police_3}
police_gun_deaths <- gun_deaths %>%
    select(police, age, race) %>%
    filter(age != "Not Available") %>%
    mutate(age = as.numeric(age)) %>%
    group_by(police)
    
ggplot(police_gun_deaths, mapping = aes(x = factor(police), y = age, shape = race, color = race)) +
    geom_point () +
    facet_wrap(.~race, ncol = 5) +
    labs(title = "Intersection of Police Involvement, Age and Race in US Gun Deaths", 
          x = "Police Involvement", 
          y = "Age (Years)") +
    scale_x_discrete(labels=c("Uninvolved", "Involved")) +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.6))
```
*I don't think that this is the best graph to represent the intersection of the three variables. Maybe I should have done percentages of race and create a pie chart, I am not sure. Can we even make a pie chart using R? This is something I shall look into!

## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

