---
title: "Exploring Gun Deaths in America"
author: Ariana Tortolani
date: 07/27/2020
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(ggplot2)
library(knitr)

# read in the data
gun_deaths <- read.csv("gun_deaths.csv")

#initial look at data
glimpse(gun_deaths)

#change month col (numbers) to month names
gun_deaths$month <- factor(gun_deaths$month, labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}

deaths_per_month <- gun_deaths %>%
    group_by(month) %>%
    count()

kable(deaths_per_month, format = "markdown",
      col.names = c("Month", "Number of Deaths"),
      caption = "Number of gun related deaths per month.")

```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
#create plot
ggplot(gun_deaths, aes(month)) +
    geom_bar(fill = "cornflowerblue") +
    scale_y_continuous(breaks = seq(0, 9000, by = 1000)) +
    
#add labels
    labs(title = "Gun Deaths per Month from 2012-2014",
         x     = "Month",
         y     = "Number of Deaths") +
    
 #add theme   
    theme_classic()
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
#create plot
    #fct_infreq(): orders by number of observations with each level (largest first)
ggplot(gun_deaths, aes(fct_infreq(intent))) +
    geom_bar(fill = "cornflowerblue") +
    scale_y_continuous(breaks = seq(0, 60000, by = 10000)) +
    
#add labels
    labs(title = "Intent of Gun Related Deaths from 2012-2014",
         x     = "Intent",
         y     = "Number of Deaths") +
    
 #add theme   
    theme_classic()
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
#create new data frame to remove NA age
gun_deaths_by_sex <- gun_deaths %>%
    filter(age != "Not Available") %>%
    mutate(age = as.numeric(age))

# generating a boxplot visualizing the data
ggplot(gun_deaths_by_sex, aes(sex, age)) +
    geom_boxplot(fill = "cornflowerblue") +
    scale_y_continuous(breaks = seq(0, 100, by = 10)) +

#add labels
    labs(title = "Gun Related Deaths by Sex from 2012-2014",
         x     = "Sex",
         y     = "Age (years)") +
    
 #add theme   
    theme_classic()


#find average age of female gun death victims and create a new data frame
female_avg_age <- gun_deaths %>%
    group_by(sex) %>%
    filter(sex == 'F') %>%
    mutate(age = as.numeric(age)) %>%
    summarize(average_age_female = mean(age, na.rm = TRUE)) # na.rm removes NA values

#print out using kable()
kable(female_avg_age, format = "markdown",
      col.names = c("Sex", "Average age of Deaths"),
      caption = "Average age of all female gun related deaths from 2012-2014")
```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
#create new data frame
white_m_hs_ed <- gun_deaths %>%
    filter(sex == "M", 
           race == "White", 
           education %in% c('HS/GED', 'Some college', 'BA+'),
           year ==2012) %>%
    count()

#print out using kable()
kable(white_m_hs_ed, format = "markdown",
      col.names = "Number of Deaths",
      caption = "Total number of white male gun related deaths with at least a high school education in 2012 ")
```


## Which season of the year has the most gun deaths?
    ANSWER: From the table printed the season with the highest number of gun deaths is Summer.

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
#create new data frame, 
    #fct_collapse: Collapse factor levels into manually defined groups
gun_deaths_season <- gun_deaths %>%
    mutate(season = fct_collapse(.f = month,
                                 Winter = c("Jan", "Feb", "Mar"),
                                 Spring = c("Apr", "May", "Jun"),
                                 Summer = c("Jul", "Aug", "Sep"),
                                 Autumn = c("Oct", "Nov", "Dec"))) %>%
    count(season) %>%  # count number of deaths per season
    arrange(desc(n))

#print out using kable()
kable(gun_deaths_season, format = "markdown",
      col.names = c("Season", "Number of Deaths"),
      caption = "Total number of gun related deaths per season")
```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
#filter data
filter_race <- gun_deaths %>%
    filter(race %in% c("White", "Black", "Hispanic"),
           intent %in% c("Suicide", "Homicide"))

#create plot
ggplot(filter_race, aes(race, fill = intent)) +
    geom_bar(position = "dodge") +
    scale_y_continuous(breaks = seq(0, 60000, by = 10000)) +
    
#add labels
    labs(title = "Number of Gun Deaths by Race and Intent from 2012-2014",
         x     = "Race",
         y     = "Number of Deaths") 
```

(hint maybe looking at percentages would be good)
```{r race_2}
#find percentage
new_filter_race<- filter_race %>%
    select(race, intent) %>%
    group_by(race, intent) %>%
    count(race) %>%
    group_by(race) %>%
    mutate(percent = 100*n / sum(n)) #new varaible to show the percentage of deaths within a race

kable(new_filter_race, format = "markdown",
      col.names = c("Race", "Intent", "Number of Deaths", "Percentage of Death (by race)"),
      caption = "Total number of gun related deaths by Race and Intent")

#create plot
ggplot(new_filter_race, aes(race, percent, fill = intent)) +
    geom_col(position = "fill") +
    
#add labels
    labs(title = "Percentage of Gun Deaths by Race from 2012-2014",
         x     = "Race",
         y     = "Percentage of Deaths") 

```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}
#filter and clean data
gun_deaths_police <- gun_deaths %>%
    select(age, police, race) %>%
    filter(age != "Not Available") %>%
    mutate(age = as.numeric(age))

# generating a boxplot visualizing the data
ggplot(gun_deaths_police, aes(police, age, group = police)) +
    geom_boxplot(fill = "cornflowerblue") +
    scale_y_continuous(breaks = seq(0, 100, by = 10)) +
    scale_x_discrete(breaks = c("0.0", "1.0"), labels = c("Uninvolved", "Involved")) +
        #it does not seem to be labeling "Uninvolved vs Involved" on the x-axis and I am unsure why.
    
#add labels
    labs(title = "Gun Related Deaths Involving Police from 2012-2014",
         x     = "Police Involvement",
         y     = "Age (years)") +

 #add theme   
    theme_classic()
```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}
# create  variables for t.test(x,y) function
police_uninvolved_age <- gun_deaths %>%
    select(police, age) %>%
    mutate(age = as.numeric(age)) %>%
    filter(police == 0)

police_involved_age <- gun_deaths %>%
    select(police, age) %>%
    mutate(age = as.numeric(age)) %>%
    filter(police == 1)

# t-test 
t.test(police_uninvolved_age$age, police_involved_age$age)

```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}
#filter data
gun_deaths_all_together <- gun_deaths %>%
    select(age, race, police) %>%
    mutate(age = as.numeric(age)) %>%
    group_by(race, police)

#create plot
ggplot(gun_deaths_all_together, aes(police, age, group = police, fill = police)) + 
            #i know the continuous scale of police on the side doesn't make much sense but I can't seem to label this how i want using scale_x_discrete
    geom_boxplot() +
    facet_wrap(~race, nrow = 1, labeller = label_wrap_gen(width = 20)) +

#add labels
    scale_y_continuous(breaks = seq(0, 100, by = 10)) +
    scale_x_discrete(breaks=c("0", "1"), labels=c("Uninvolved", "Involved")) + #still not showing labels when i run this

#add labels
    labs(title = "Gun Related Deaths from 2012-2014",
         x     = "Police Involvement",
         y     = "Age (years)") 
```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

