---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
gun_deaths <- read_csv("gun_deaths.csv") # read in the data

```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
library(knitr) #load knitr, required for kable()

gun_deaths_by_month <- gun_deaths %>%
    select(month, year) %>%
    group_by(month) %>% #Orders table by month
    count(year, name = "deaths_month") %>% #Enumerates gun deaths
    mutate(abMonth = month.abb[month], sd = sd(deaths_month)) %>% #Adds column of abbreviated month name and standard error
    ungroup()

#Creates table with year, month and deaths
gun_deaths_by_month %>%
    select(year, abMonth, deaths_month) %>%
    arrange(year) %>%
    rename("Year |" = "year", "Month |" = "abMonth", "American Gun Deaths" = "deaths_month") %>%
    kable(format = "markdown")
```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
#Makes abbreviated months leveled factors for proper display via ggplot
gun_deaths_by_month$abMonth <- factor(gun_deaths_by_month$abMonth, 
                                      levels = c(month.abb[12:1]))
#Plots data of deaths by month
set.seed(3)
gun_deaths_by_month  %>% 
    ggplot(mapping = aes(x = abMonth, y = deaths_month)) +
    geom_bar(stat = "summary", fun = "mean", alpha = 0.7, width = 0.4) +
    geom_errorbar(stat = "summary", 
                  mapping = aes(ymin = deaths_month - sd, 
                                ymax = deaths_month + sd, width = 0.2)) +
    scale_y_continuous() + coord_flip() +
    xlab("Month") + ylab("American Gun Deaths\n(average, standard deviation)") +
    ggtitle("American Gun Deaths by Month (2012-2014)")
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
#Selects and enumerates death by intent, trimmed unavailable datum
gun_deaths_by_intent <- gun_deaths %>%
    select(intent) %>%
    count(intent, name = "deaths_intent") %>%
    filter(intent != "Not Available") %>%
    arrange(desc(deaths_intent))

#Plots data of deaths by determined intent
gun_deaths_by_intent %>%
ggplot(mapping = aes(x = reorder(intent, -deaths_intent), y = deaths_intent)) +
    geom_bar(stat = "identity", alpha = 0.7, width = 0.7) +
    xlab("Determined Intent of Gun Death") + ylab("Number of Gun Deaths") +
    ggtitle("American Gun Deaths by Intent (2012-2014)")
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
#Selects age and sex variables from gun_deaths
gun_deaths_by_age_sex <- gun_deaths %>%
    select(age, sex) %>%
    filter(age != "Not Available") %>% 
    transform(age = as.numeric(age)) %>% 
    group_by(sex) %>%
    mutate(mean_age = mean(as.numeric(age))) %>%
    add_count(sex, name = "deaths_sex") %>%
    ungroup()

#Boxplot of age vs sex
gun_deaths_by_age_sex %>% 
    ggplot(mapping = aes(x = sex, y = age, color = sex)) +
    geom_boxplot(fill = NA, outlier.shape = NA, width = 0.595) + 
    geom_point(size = 0, position = position_jitter(width = 0.3), alpha = 0.1) +
    scale_color_brewer(palette = "Set2") +
    xlab("Sex of Victims") + ylab("Age of Victims") +
    ggtitle("Sex and Age of Gun Death Victims in America (2012-2014)")
    

gun_deaths_by_age_sex[2,2:3]
```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
#Enumerating number of deaths of white males with at least a high school education in 2012
gun_deaths %>%
    select(year, sex, race, education) %>% 
    filter(sex == "M", year == 2012, education != "Less than HS", race == "White") %>% 
    count(education, name = "white male American gun victims in 2012") %>%
    bind_rows(summarise_all(., funs(if(is.numeric(.)) sum(.) else "Total"))) %>% #Adds row for Total number of deaths
    kable(format = "markdown")

```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
#Adds season column to gun_deaths_by_month using cut() function
gun_deaths_by_month$season <- cut(gun_deaths_by_month$month, c(0,3.5,6.5,9.5,13), 
                                  c("Winter", "Spring", "Summer", "Fall"))

#Adds deaths per season and year
gun_deaths_by_month <- gun_deaths_by_month %>% 
    group_by(season, year) %>%
    mutate(deaths_season = sum(deaths_month)) %>%
    mutate(avg_deaths_season = mean(deaths_season)) %>% 
    ungroup()
 
#Produce table to look at values   
gun_deaths_by_month %>%
    select(year, season, deaths_season) %>%
    distinct() %>%
    arrange(year)

#Generates barplot of deaths
gun_deaths_by_month %>%
    arrange(desc(year)) %>%
    distinct() %>% 
    ggplot(mapping = aes(x = season, y = deaths_month, fill = as.character(year))) +
    geom_bar(stat = "identity", alpha = 0.7) + 
    scale_y_continuous(limits = c(0,29000)) +
    scale_fill_brewer(palette = "Set2") +
    xlab("Season") + ylab("Gun Death Victims") +
    ggtitle("Season and Year of Gun Deaths in America (2012-2014)") +
    theme(legend.title = element_blank(), legend.position = c(0.5, 0.93),
          legend.background = element_blank(), legend.direction = "horizontal")
```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}

```

(hint maybe looking at percentages would be good)
```{r race_2}

```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}

```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}


```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

