---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 
    


## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
install.packages("tidyverse")
library(tidyverse)    # load tidyverse packages, including ggplot2
library(ggplot2)
# read in the data
gunDeaths <-read.csv(file="gun_deaths.csv", head=TRUE, sep=",")
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
library(knitr)
myTable <-kable(gunDeaths, format="markdown" )
head(myTable)
```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
ggplot(gunDeaths, aes(month))+
    geom_bar() +
    scale_x_continuous( "Month",
     breaks = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12),
     labels = c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'June','July','Aug','Sept', 'Oct', 'Nov', 'Dec'))+
    labs(title="Gun Deaths in 2012", y="Death Count")
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
# ggplot(gunDeaths, aes(month, fill=intent))+
#     geom_bar(position="dodge")

gD_Intent<- gunDeaths %>%
    group_by(intent) %>%
    count() 

gD_Intent$intent<- factor(gD_Intent$intent, 
                levels=c("Suicide", "Homicide", "Accidental", "Undetermined", "Not Available")) 

ggplot(gD_Intent, aes(x=intent, y= n, fill=intent))+
    geom_bar(stat="identity")+
    labs(title="Gun Deaths by Intentions in 2012", x=element_blank(), y="Death Count", fill="Reported Intentions")
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
gunDeaths$age<- as.numeric(gunDeaths$age)  # had to add this because my argument wasn't numeric
ageFemale <- gunDeaths %>%
    select(sex, age) %>%
    filter(sex == "F")
avgAgeFemale<- mean(ageFemale$age)
#avgAgeFemale

ggplot(gunDeaths, aes(sex, age, fill=sex))+
    geom_boxplot()+
    annotate("text", x=1, y=80, label=paste("Avg F age :", avgAgeFemale))+
    labs(title="Ages of gun-caused deaths by gender", x=element_blank(), y="Age", fill="Gender")
   

```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}
wMales<- gunDeaths %>%
    select(sex, race, education) %>%
    filter(education %in% c("HS/GED", "Some college", "BA+"), race== "White", sex=="M") %>%
    count()
wMales
#46227 were killed by guns in 2012.
```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
gunDeaths$month <- factor(gunDeaths$month, labels= c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
Winter<- gunDeaths %>%
    select(month) %>%
    filter(month %in% c("Jan", "Feb", "Mar")) %>%
    count() 
Spring<- gunDeaths %>%
    select(month) %>%
    filter(month %in% c("Apr", "May", "Jun")) %>%
    count() 
Summer<- gunDeaths %>%
    select(month) %>%
    filter(month %in% c("Jul", "Aug", "Sep")) %>%
    count() 
Fall<- gunDeaths %>%
    select(month) %>%
    filter(month %in% c("Oct", "Nov", "Dec")) %>%
    count() 
Winter
Spring
Summer
Fall
max(Winter, Spring, Summer, Fall)
#Summer has max deaths = 26280
```




## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

