---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(dplyr)
library(magrittr)
library(knitr)

# read in the data
gun_death<- read_csv("gun_deaths.csv")

```
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)

#glimps
glimpse(gun_death)
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"

```{r month, warning=FALSE}
month_death<- gun_death %>%
    group_by(month) %>%
    count(month)

kable(month_death, format="markdown", row.names= NA, col.names= c("Month", "Number of Deaths"), caption= "Monthly gun-related deaths in 2012, data retrieved from https://github.com/fivethirtyeight/guns-data")
```

### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}
#change variable month to contain abbs of months
gun_death$month <- factor(gun_death$month, labels= c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

#create a data frame that has the count of death for each month 
monthly_death<- gun_death %>%
    group_by(month) %>%
    count()

#plot   
ggplot(monthly_death, aes(x=month, y=n))+
    geom_bar(stat="identity", fill="red", alpha=0.7)+
    scale_y_continuous(limit=c(0,12000))+
    labs(title="Montly Gun Deaths in 2012", x=element_blank(), y="Number of deaths")+
    theme_bw()
```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}
#create a data frame with death number associated with intent type
gun_death_intent<- gun_death %>%
    group_by(intent) %>%
    count() 

#A way to level the intent variable in order to order the bins, not the best way if there are more variables
gun_death_intent$intent<- factor(gun_death_intent$intent, 
                                 levels=c("Suicide", "Homicide", "Accidental", "Undetermined", "Not Available"))     

#plot
ggplot(gun_death_intent, aes(x=intent, y=n, fill=intent))+
    geom_bar(stat="identity", alpha=0.9)+
    scale_y_continuous(limit=c(0,64000))+
    labs(title="Number of Gun Deaths by Different Intentions in 2012", x=element_blank(), y="Number of deaths", fill="Reported Intentions")+
    theme_bw()
```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}
##make age variable numeric
gun_death$age<- as.numeric(gun_death$age)

#mean of female victims
female_age<- gun_death %>%
    select(age, sex) %>%
    filter(sex=="F") 

avg<- mean(female_age$age, na.rm=TRUE)  
avg<- round(avg, digits=2)

#plot
ggplot(gun_death, aes(x=sex, y=age, fill=sex))+
    geom_boxplot()+
    annotate("text", x=1, y=70, label=paste("The average age\n of female victim is", avg))+
    labs(title="Ages of Gun Violence Victims in 2012, By Gender", x=element_blank(), y="Age", fill="Gender")+
    theme_bw()
```

## How many white males with at least a high school education were killed by guns in 2012?

-The answer is 46227 

```{r education}
answer<- gun_death %>%
    select(education, race, sex) %>%
    filter(education %in% c("HS/GED", "Some college", "BA+"), race== "White", sex=="M") %>%
    count()
answer
```

## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

-Summer has the most gun deaths.

```{r season}
#calculate death count in each season
winter<- gun_death %>%
    select(month) %>%
    filter(month %in% c("Jan", "Feb", "Mar")) %>%
    count() 

spring<- gun_death %>%
    select(month) %>%
    filter(month %in% c("Apr", "May", "Jun")) %>%
    count() 

summer<- gun_death %>%
    select(month) %>%
    filter(month %in% c("Jul", "Aug", "Sep")) %>%
    count() 

fall<- gun_death %>%
    select(month) %>%
    filter(month %in% c("Oct", "Nov", "Dec")) %>%
    count() 

max(winter, spring, summer, fall)

```
# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}
#data frame of gun deaths intent 
gun_death_race3<- gun_death %>%
    filter(race %in% c("White", "Black", "Hispanic"), intent %in% c("Suicide", "Homicide")) 

#order race
gun_death_race3$race<- factor(gun_death_race3$race, 
                                 levels=c("White", "Black", "Hispanic")) 

#plot
ggplot(gun_death_race3, aes(x=race, fill=intent))+
    geom_bar(position="stack", alpha=0.9)+
    labs(title="Gun Death Intentions Comparisons Among Three Populations", x="Race", y="Number of deaths", fill="Reported Intentions")+
    theme_bw()
```

(hint maybe looking at percentages would be good)
```{r race_2}

#add a percent column summarizing the info
gun_death_percent<- gun_death_race3 %>%
    group_by(race, intent) %>%
    count() %>%
    rename(death=n) %>%
    group_by(race) %>%
    mutate(total=sum(death), percent=death/total*100)

#plot
ggplot(gun_death_percent, aes(x=race, y=percent, fill=intent))+
    geom_bar(stat="identity", position="stack", alpha=0.9)+
    labs(title="Proportions of Reported Intentions of Gun Deaths Among Three Populations", x="Race", y="Percentage(%)", fill="Reported Intentions")+
    theme_bw()
```

### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}
#data frame
gun_death_police<- gun_death %>%
    select(sex, age, police) 

#make police variable factor
gun_death_police$police<- factor(gun_death_police$police)

#Are police-involved gun deaths significantly different from other gun deaths?
gun_death_police %>%
    group_by(police) %>%
    count()
#plot
ggplot(gun_death_police, aes(x=police))+
    geom_bar(fill="blue", alpha=0.7)+
    labs(title="The Majority of Gun Deaths in 2012 Did Not Involve A Police", x="Police Involvement\n(No=0, Yes=1)", y="Number of deaths")+
    theme_bw()
```

## Comparing the mean ages of gun death victims with or without police involvement, which is younger?

- A t test comparing the means of age among gun deaths w/o police involvement shows that there is a statistically significant difference the age means. 
- A boxplot below to visualize that the average age of the vicimts of police-involved gun deaths is younger. 

```{r police_&_age}
#make age variable numeric
gun_death_police$age<- as.numeric(gun_death_police$age)

#t test
t.test(gun_death_police$age ~ gun_death_police$police)

#plot
ggplot(gun_death_police, aes(x=police, y=age, fill=police))+
    geom_boxplot(alpha=0.9)+
    labs(title="Age Differences of Gun Deaths w/o Police Involvement", x="Police Involvement\n(No=0, Yes=1)", y="Age")+
    theme_bw()
```

## Are victims of police-involved gun death more likely to be male or female?

- The visualization shows a greater proportion of police-involved victims are male compared to gun death that did not invovle a police. 
- A Chi-square test shows statistical sigficance of this comparison.

```{r police_&_sex}
police_sex_percent<- gun_death_police %>%
    group_by(sex, police) %>%
    count() %>%
    group_by(sex) %>%
    mutate(total=sum(n), percent=n/total)

ggplot(police_sex_percent, aes(x=police, y=percent, fill=sex))+
    geom_bar(stat="identity", position="fill", alpha=0.7)+
    labs(title="Police Involvement in Gun Deaths By Sex", x="Police Involvement\n(No=0, Yes=1)", y="Percent(%)")+
    theme_bw()
```

```{r police_2}
chisq.test(gun_death_police$police, gun_death_police$sex, correct=FALSE)

```

## Age, gender, and police involvement of gun deaths in 2012

```{r police_3}
ggplot(gun_death_police, aes(x=police, y=age, fill=sex))+
    geom_boxplot(alpha=0.9)+
    labs(title="Age Differences of Gun Deaths w/o Police Involvement By Gender", x="Police Involvement\n(No=0, Yes=1)", y="Age")+
    theme_bw()
```

## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

