---
title: "Exploring Gun Deaths in America"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(knitr) # load knitr package so that I can use kable() function

# read in data and define as gun_deaths
gun_deaths <- read.csv("gun_deaths.csv")

```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}

# generate new dataframe for this task
gun_deaths_summary1 <- gun_deaths %>%
    select(month) %>%
    group_by(month) %>%
    count(month) %>%
    rename(deaths_per_month = n)

# print results as a table using kable()
kable(gun_deaths_summary1, format = "markdown", align = 'c', col.names = c("Month", "Deaths per Month"))


```

**If we further want to divide the number of gun deaths per month for each given year:**

```{r month_2, warning=FALSE}

# generate new dataframe, almost the same as above but also selecting and grouping by year
gun_deaths_summary2 <- gun_deaths %>%
    select(year, month) %>%
    group_by(year, month) %>%
    count(month) %>%
    rename(deaths_per_month = n)

# print results as a table using kable()
kable(gun_deaths_summary2, format = "markdown", align = 'c', col.names = c("Year", "Month", "Deaths per Month"))


```



### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}

# generate bar graph using the dataframe generated above
ggplot(data = gun_deaths_summary1, aes(x = factor(month), y = deaths_per_month)) + # make sure month is plotted as factor so that we can convert to abbreviated month name
    geom_col() +
    scale_x_discrete(labels=month.abb) + # change x-axis labels to abbreviated month names instead of numbers
    labs(title = "Number of Gun Deaths for each Month of the Year", subtitle = "Based on Data from 2012 - 2014", x = "Month", y = "Number of Deaths") + # adjust labels
    theme_classic()   # add a theme (I like to use the pre-set theme theme_classic())


```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}

# generate new dataframe looking at the number of gun deaths associated with ech type of intent
gun_deaths_intent <- gun_deaths %>%
    select(year, month, intent) %>%
    group_by(intent) %>%
    count() %>%
    rename(number_of_deaths = n)

# print results as a table using kable()
kable(gun_deaths_intent, format = "markdown", align = 'c', col.names = c("Intent", "Number of Deaths"))
```

**We can use this output to sort the x-axis accordingly:**

```{r intent_2}

# sort the x-axis based on table above
gun_deaths_intent$intent <- factor(gun_deaths_intent$intent, levels = c("Suicide", "Homicide", "Accidental", "Undetermined", "Not Available"))

# generate bar graph showing the number of gun deaths associated with each type of intent from highest to lowest value
ggplot(gun_deaths_intent, aes(x = intent, y = number_of_deaths, color = intent, fill = intent)) +
    geom_col() +
    labs(title = "Number of Gun Deaths based on Intent", subtitle = "Based on Data from 2012 - 2014", x = "Intent", y = "Number of Deaths") + #adjust labels
    theme_classic() + # set theme
    theme(legend.position = "none") # remove legend

```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}

# generate new dataframe looking at the age of gun death victims by sex
gun_deaths_sex <- gun_deaths %>%
    select(sex, age) %>%
    group_by(sex) 

# change data type of age column to integer
gun_deaths_sex$age <- as.integer(gun_deaths_sex$age)

# generate boxplot 
ggplot(gun_deaths_sex) +
    geom_boxplot(aes(sex, age, fill = sex)) +
    labs(title = "Age of Gun Death Victims by Sex", x = "Sex", y = "Age") + # adjust labels
    theme_bw() + # set theme
    theme(legend.position = "none") # remove legend
    

# find the average age of female gun death victims
gun_deaths_average_female <- gun_deaths_sex %>%
    filter(sex == "F") %>%
    summarize(average_age = mean(age, na.rm = TRUE)) # na.rm allows to 'remove' NA values

head(gun_deaths_average_female)

```


**The average female gun death victim age was ~43.7 years.**

## How many white males with at least a high school education were killed by guns in 2012?

```{r education}

# generate new dataframe looking at how many white males with at least a high school education were killed by guns in 2012
gun_deaths_education <- gun_deaths %>%
    select(sex, education, year, race) %>%
    filter(sex == "M", year == 2012, race == "White", education %in% c('HS/GED', 'Some college', 'BA+')) %>% # I believe that this should include all education levels with at least a high school education
    count()

# print the number of white males killed as kable() output
kable(gun_deaths_education, format = "markdown", align = 'c', col.names = c("Number of White Males killed"))

```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}

# generate new dataframe to look at the season of the year with the most gun deaths
gun_deaths_season <- gun_deaths %>%
    select(month) 

# convert the 'month' variable to a factor
gun_deaths_season$month <- as.factor(gun_deaths_season$month)

# convert from months to season
gun_deaths_season%>%
    mutate(season = fct_collapse(.f = month,
      Spring = c("4", "5", "6"),
      Summer = c("7", "8", "9"),
      Autumn = c("10", "11", "12"),
      Winter = c("1", "2", "3"))) %>%
    count(season) %>% # count number of deaths per season
    arrange(desc(n)) # arrange in descending order so we can see which season has the most deaths





```

**As seen by the table above, the season with the most gun deaths is summer**


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}

# generate new dataframe looking at number of gun deaths based on race and intent
gun_deaths_race <- gun_deaths %>%
    select(intent, race) %>%
    filter(intent %in% c('Suicide', 'Homicide')) %>%
    filter(race %in% c('White', 'Black', 'Hispanic')) %>%
    group_by(race, intent) %>%
    count(race)

# generate bar chart to visualize the dataframe generated above
ggplot(gun_deaths_race, aes(x = gun_deaths_race$race, y = n, fill = intent)) +
    geom_col(position = "dodge", color = "black") +
    labs(title = "Number of Gun Deaths by Race and Intent", x = "Race", y = "Number of Deaths") +
    scale_fill_discrete(name = "Intent") + # adjust labels
    theme_classic() # set theme


```

(hint maybe looking at percentages would be good)
```{r race_2}

# generate new dataframe to calculate the percentage of deaths based on race and intent
gun_deaths_race_2 <- gun_deaths_race %>%
    ungroup() %>%
    group_by(race) %>%
    mutate(percentage = 100*n / sum(n)) # generate new variable that shows the percentage

# print the new dataframe as kable() output
kable(gun_deaths_race_2, format = "markdown", align = 'c', col.names = c("Race", "Intent", "Number of Deaths", "Percentage (by Race)"))

# generate bar chart to visualize the percentage of deaths based on race and intent
ggplot(gun_deaths_race_2, aes(x = gun_deaths_race$race, y = percentage, fill = intent)) +
    geom_col(position = "dodge", color = "black") +
    labs(title = "Percentage of Gun Deaths by Intent", x = "Race", y = "Percentage of Deaths by Intent") + # adjust labels
    scale_fill_discrete(name = "Intent") + # this allows to rename the legend title
    theme_classic() # set theme
```


**Looking at the percentage of gun deaths by intent, we find that for Black and Hispanic, the majority of gun death victims died as a result of homicide (85% and 64% respectively), whereas for white people, the majority of gun death victims died as a result of suicide (86%).**

### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}

# change the format of the age and police variables of the gun_deaths dataframe
gun_deaths$age <- as.integer(gun_deaths$age)
gun_deaths$police <- as.factor(gun_deaths$police)

# generate new dataframe looking at number of gun deaths based on police involvement and age
gun_deaths_police_age <- gun_deaths %>%
    select(police, age) %>%
    group_by(police) %>%
    count(age)
    
# generate boxplot visualizing the number of gun deaths based on police involvement and age
ggplot(gun_deaths_police_age, aes(x = police, y = age, fill = police)) +
    geom_boxplot() +
    scale_x_discrete(breaks = c("0", "1"), labels = c("No Police Involvement", "Police Involvement")) + # rename x-axis labels
    labs(title = "Age of Gun Death Victims based on Police Involvement", x = "", y = "Age") + # adjust labels
    theme_classic() + # set theme 
    theme(legend.position = "none") # remove legend


```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}

# I want to test for significance by doing a t-test

# generate new variables that can be used in t.test(x,y) function
police_not_involved_age <- gun_deaths %>%
    select(police, age) %>%
    filter(police == 0)

police_involved_age <- gun_deaths %>%
    select(police, age) %>%
    filter(police == 1)

# do t-test using the newly generated variables
t.test(police_not_involved_age$age, police_involved_age$age)



```

#### The t-test suggests that there is a significant difference in the age of gun death victims based on whether police was involved or not. 

```{r police_2}

# generate new dataframe looking at number of gun deaths based on police involvement and race
gun_deaths_police_race <- gun_deaths %>%
    select(police, race) %>%
    group_by(race, police) %>%
    count(race) %>%
    arrange(race)

# generate bar graph visualizing number of gun deaths based on police involvement and race
ggplot(gun_deaths_police_race, aes(x = race, y = n, fill = police)) +
    geom_col(position = "dodge", color = "black") +
    scale_x_discrete(breaks = c("Asian/Pacific Islander", "Black", "Hispanic", "Not Availabletive American/Not Availabletive Alaskan", "White"), labels = c("Asian/Pacific Islander", "Black", "Hispanic", "NA", "White")) + # adjust labels on x-axis
    labs(title = "Number of Deaths based on Police Involvement by Race", x = "Race", y = "Number of Deaths") + # adjust other labels 
    scale_fill_discrete(labels = c("No Police Involvement", "Police Involvement")) + # adjust legend text
    theme_classic() + # set theme
    theme(legend.title = element_blank()) # remove legend title



```



Finally, all together now:
```{r police_3}

# generating a new dataframe, looking at age of gun death victims based on race and police involvement
gun_deaths_police_age_race <- gun_deaths %>%
    select(police, age, race) %>%
    group_by(race, police) 
    
# generate boxplot visualizing age of gun death victims based on race and police involvement
ggplot(gun_deaths_police_age_race, aes(x = race, y = age, fill = police)) +
    geom_boxplot(outlier.shape = NA) +
    scale_x_discrete(breaks = c("Asian/Pacific Islander", "Black", "Hispanic", "Not Availabletive American/Not Availabletive Alaskan", "White"), labels = c("Asian/Pacific Islander", "Black", "Hispanic", "NA", "White")) + # adjust labels on x-axis
    labs(title = "Age of Gun Death Victims based on Race and Police Involvement", x = "Race", y = "Age") + #adjust other labels
    scale_fill_discrete(labels = c("No Police Involvement", "Police Involvement")) + # adjust legend text
    theme_classic() + # set theme
    theme(legend.title = element_blank()) # remove legend title

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

