---
title: "Exploring Gun Deaths in America"
author: "Nicholas Elliott"
output: github_document
---

## Objectives
* To explore gun death data curated by FiveThirtyEight [original data available here](https://github.com/fivethirtyeight/guns-data) by
    * Month
    * Intent
    * Age and sex
    * Education
    * Season
    * Intent by race
    * Police-relatedness 

## Read in the data

The gun_deaths.csv is provided in the repository. 
```{r load-data, message=FALSE, warning=FALSE}
library(tidyverse)    # load tidyverse packages, including ggplot2
library(knitr)
# read in the data
gun_deaths <- read_csv("gun_deaths.csv", col_types = "dccdcdcdcc")
head(gun_deaths)
```

## Generating a data frame that summarizes the number of gun deaths per month and printing using `kable()`. 

Knitr is part of the knitr package. Allows you to change column names and add captions and make pretty tables in your knitted document. Hint: set the argument format = "markdown"


```{r month, warning=FALSE}
deathsPermonth <- gun_deaths %>%
    group_by(month) %>%
    count(month, name = "deaths")

deathsPermonth_df <- data.frame(deathsPermonth)

kable(deathsPermonth_df, "markdown")

```


### Generating a bar chart with human-readable labels on the x-axis. That is, each month should be labeled "Jan", "Feb", "Mar" (full or abbreviated month names are fine), not `1`, `2`, `3`.

```{r month_chart}

deathsPermonth_df %>%
ggplot(aes(x = reorder(month, + deaths), y = deaths)) +
    geom_bar(stat = "identity", color = "black", fill = "blue") +
    scale_x_discrete(labels = c("Jan", "Oct", "Nov", "Dec", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep")) +
    xlab("month")

```

## Generating a bar chart that identifies the number of gun deaths associated with each type of intent cause of death. The bars should be sorted from highest to lowest values.

```{r intent}

gun_deaths %>%
    group_by(intent) %>%
    count(intent, name = "n") %>%
ggplot(aes(x = reorder(intent, - n), y = n)) +
    geom_bar(stat = "identity", color = "black", fill = "green") +
    ylab("deaths") +
    xlab("intent")

```

## Generating a boxplot visualizing the age of gun death victims, by sex. Print the average age of female gun death victims.

```{r age_&_sex}

gun_deaths %>%
    group_by(age, sex) %>%
ggplot(aes(sex, age, fill = sex)) +
    geom_boxplot()

avg_age <- gun_deaths %>%
    group_by(age, sex) %>%
    filter(sex == "F") %>%
    count(age, name = "age_count")
    
avg_age_df <- data.frame(avg_age)
avg_age_df
apply(avg_age_df, 2, mean)
    
```


## How many white males with at least a high school education were killed by guns in 2012?

```{r education}

white_M_gunDeaths <- gun_deaths %>%
    select(year, sex, race, education) %>%
    filter(year == 2012, sex == "M", race == "White", 
           education %in% c("BA+", "HS/GED", "Some college")) %>%
    count(education, name = "deaths") %>%
    summarize(HS_educated_white_males_killed_2012 = sum(deaths))

kable(white_M_gunDeaths, "markdown")

```


## Which season of the year has the most gun deaths?

Assume that:

* Winter = January-March
* Spring = April-June
* Summer = July-September
* Fall = October-December

```{r season}
#Most deaths are in Summer. Is there a quicker way to do this? Seems like a lot for a little.

Winter <- gun_deaths %>%
    filter(month %in% c("1", "2", "3")) %>%
    count(month, name = "winter_deaths") %>%
    summarize(Deaths = sum(winter_deaths))

Spring <- gun_deaths %>%
    filter(month %in% c("4", "5", "6")) %>%
    count(month, name = "spring_deaths") %>%
    summarize(Deaths = sum(spring_deaths))

Summer <- gun_deaths %>%
    filter(month %in% c("7", "8", "9")) %>%
    count(month, name = "summer_deaths") %>%
    summarize(Deaths = sum(summer_deaths))

Fall <- gun_deaths %>%
    filter(month %in% c("10", "11", "12")) %>%
    count(month, name = "fall_deaths") %>%
    summarize(Deaths = sum(fall_deaths))

Winter_df <- data.frame(Winter, row.names = "Winter")
Spring_df <- data.frame(Spring, row.names = "Spring")
Summer_df <- data.frame(Summer, row.names = "Summer")
Fall_df <- data.frame(Fall, row.names = "Fall")

Seasons <- rbind(Winter_df, Spring_df, Summer_df, Fall_df)

kable(Seasons, "markdown")

```


# These are more open ended questions from here on out, you could look at making multiple summaries or types of plots. 

### Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

```{r race}

intent_deaths <- gun_deaths %>%
    group_by(race, intent) %>%
    filter(intent %in% c("Homicide", "Suicide"), race %in% c("White", "Black", "Hispanic")) %>%
    count(intent, name  = "H_S_Deaths")

ggplot(intent_deaths, aes(x = race, y = H_S_Deaths, fill = intent)) +
    geom_col(position = "dodge") +
    ylab("Gun Deaths") +
    xlab(element_blank())

```

(hint maybe looking at percentages would be good)
```{r race_2}
intent_deaths <- gun_deaths %>%
    group_by(race, intent) %>%
    filter(intent %in% c("Homicide", "Suicide"), race %in% c("White", "Black", "Hispanic")) %>%
    count(intent, name  = "H_S_Deaths") %>%


    


```


### Are police-involved gun deaths significantly different from other gun deaths? Assess the relationship between police involvement and age, police involvement and race, and the intersection of all three variables.

```{r police}

```

Think about performing stats such as a t.test to see if differences you are plotting are statistically significant

```{r stats?}


```


```{r police_2}

```

```{r stats_2?}


```


Finally, all together now:
```{r police_3}

```


## Session info

```{r}
# always good to have this for reproducibility purposes
devtools::session_info()
```

